---
title: "Categorical Data"
subtitle: "Data Visualization"
author: "Johan Larsson"
institute: "The Department of Statistics, Lund University"
date: ""
bibliography: "assets/references.bib"
output:
  xaringan::moon_reader:
    css: [default, "assets/styles.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      titleSlideClass: [middle, center]
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dev = "png",
  fig.width = 4.8,
  fig.height = 4.1,
  fig.align = "center",
  fig.retina = 3,
  dev.args = list(pointsize = 16),
  crop = TRUE
)

library(tidyverse)
library(kableExtra)

kable <- function(x, ...) {
  knitr::kable(x, escape = FALSE, linesep = "", booktabs = TRUE)
}

library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib("assets/references.bib")

if (!interactive()) {
  theme_set(theme_grey(base_size = 16))
}

# xaringanExtra::use_search(show_icon = TRUE)

options(digits = 3)
```

## Visualizing Categorical Data

usually, the real aim is to visualize **proportions**

.pull-left[
### Example

happiness data from the US in 2006

```{r}
library(tidyverse)
data(
  "happiness",
  package = "wooldridge"
)

# see slide source code for 
# data wrangling code
```

]

```{r, echo = FALSE}
happiness <- 
  as_tibble(happiness) %>%
  mutate(
    happy = factor(
      happy, 
      levels = c("not too happy", "pretty happy", "very happy" ),
      ordered = TRUE
    ),
    gender = recode( female, "1" = "female", "0" = "male" )
  ) %>% filter(y06 == 1)
```

--

.pull-right[
### Bar Charts

simple and readable

```{r, fig.height = 3.5}
ggplot(happiness, aes(happy)) +
  geom_bar() #<<
```
]

---

## Grouped Bar Charts

good when counts are what matters or when you have small data sets

```{r, fig.width = 9, fig.height = 5}
ggplot(happiness, aes(gender, fill = happy)) +
  geom_bar(position = "dodge", col = "black") #<<
```

---

## Stacked Bar Charts

compact way to visualize counts, but seldom preferable over grouped bar chart

```{r, fig.width = 7, fig.height = 5}
ggplot(happiness, aes(gender, fill = happy)) +
  geom_bar(col = "black") # position = "stack" is the default #<<
```

---

## Proportional Grouped Bar Charts

good default choice for bar charts (when you have enough data)

but need to summarize data before plotting!

```{r, fig.width = 7, fig.height = 3.5}
happiness_props <- group_by(happiness, gender, happy) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n))

ggplot(happiness_props, aes(gender, prop, fill = happy)) +
  geom_col(position = "dodge", col = "black")      #<<
```

---

## Proportional Stacked Bar Charts

compact way to visualize proportions

slightly more intuitive than grouped proportional bar charts, but harder to
read
```{r, fig.width = 6}
ggplot(happiness, aes(gender, fill = happy)) +
  geom_bar(position = "fill", col = "black") + #<<
  labs(y = "proportion")
```

---

## Mosaic Plots

a type of stacked bar chart that maps category size to
the width of bars

need to use another package for this:
[ggmosaic](https://CRAN.R-project.org/package=ggmosaic)

```{r, fig.width = 6}
library(ggmosaic)
ggplot(happiness) +
  geom_mosaic(aes(product(happy, gender), fill = happy)) + #<<
  guides(fill = "none") # remove redundant ink
```

---

## Waffle Plots

suitable when there are large differences between categories

again need to use a new package:
[waffle](https://CRAN.R-project.org/package=waffle)

```{r, echo = FALSE, fig.width = 7, fig.height = 5, fig.cap = "A waffle plot of the happiness data. Every square represents approximately 10 people. The code is quite complicated (see the source code)."}
# devtools::install_github("hrbrmstr/waffle")
library(waffle)

happiness_counts <-
  group_by(happiness, gender) %>%
  count(happy) %>%
  mutate(n = round(n / 10))

ggplot(happiness_counts, aes(fill = happy, values = n)) +
  geom_waffle(color = "white", size = 1.125, n_rows = 10, flip = TRUE) +
  facet_wrap("gender", nrow = 1) +
  coord_equal() +
  theme_minimal(base_size = 16) +
  theme_enhance_waffle()
```


---

## What About Pie Charts?

okay in some instances (2-3 categories), but avoid as a thumb rule

.pull-left[
```{r fig.width = 4, fig.height = 5, echo = FALSE}
library(scales)

happiness_bins <-
  happiness %>%
  count(workstat) 

ggplot(happiness_bins, aes(x = "", y = n, fill = workstat)) +
  geom_col(width = 1, col = 1) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Accent") +
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    legend.pos = "bottom",
    legend.title = element_blank()
  ) +
  guides(fill = guide_legend(ncol = 2))
```
]

.pull-right[
```{r fig.height = 5, echo = FALSE}
ggplot(happiness_bins, aes(workstat, n, fill = workstat)) +
  geom_col(col = "black", show.legend = FALSE) +
  scale_fill_brewer(palette = "Accent") +
  labs(x = NULL, y = "count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
]

---

## Pay Attention to Mappings

choice of mappings is critical when dealing with categorical data!

```{r, fig.width = 7, fig.height = 4}
ggplot(happiness, aes(happy, fill = gender)) + #<<
  geom_bar(position = "fill", col = "black") +
  ylab("Proportion")
```

---

## Ordering

if variable is ordinal: just format it as an ordered factor

if variable is **not** ordinal: consider ordering it manually!

```{r, fig.width = 10, fig.height = 3.5}
# no ordering
happiness %>%
  count(region) %>%
  ggplot(aes(region, n)) +
  geom_col() +
  guides(x = guide_axis(n.dodge = 2)) # avoid label overlap
```

---

## Ordering

if variable is ordinal: just format it as an ordered factor

if variable is **not** ordinal: consider ordering it manually!

```{r, fig.width = 10, fig.height = 3.5}
happiness %>%
  count(region) %>%
  mutate(region = fct_reorder(region, n, .desc = TRUE)) %>% #<<
  ggplot(aes(region, n)) +
  geom_col() +
  guides(x = guide_axis(n.dodge = 2)) # avoid label overlap
```

