---
title: "Instructions for the Assignments"
author: "Johan Larsson"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: tango
    global_numbering: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

For each assignment, as well as for the project, you need to hand in a `.pdf`
report. This report should be a written response to the tasks given in the
assignment. It should respond to all the questions in the assignment in a
**clear** and **succinct** manner.

## Workflow: Two Approaches

There are two approaches you can take when you write your report. The first of
these is to use R Markdown; the second is to write your report in a WYSIWYG
(what you see is what you get) software such as Libre Office, Google Docs, or
Microsoft Word.

### R Markdown

R Markdown represents a way to write reproducible reports that can be rendered
in a wide variety of formats and which allows you to interleave R code with
text.

*Pros:*

* You don't need to bother with formatting details. R Markdown, pandoc, and
  LaTeX takes care of it all for you.
* It's very convenient to modify figures or change image sizes. (Simply
  change `fig.width` and `fig.height` in the header).
* Code that you include is automatically formatted and receives pretty
  syntax highlighting.
* Your document becomes reproducible by someone else if you just give them
  the `.Rmd` file (and sometimes some other additional auxiliary files). You
  are even encouraged to submit your R Markdown file along with your submission
  for the assignments and project, so that your peers can reproduce your work
  (if they are interested in doing so).

*Cons:*

* You need to learn Markdown syntax and R Markdown code chunks, which will
  require an initial time investment. If you don't expect
  to write these kinds of reports or presentations in the future or will be
  working with people who will not use R Markdown, this investment may not
  pay off.
* There are sometimes technical difficulties in getting R Markdown working,
  especially for Windows machines^[This is typically not a problem since you
  can export your document to a Word document instead of a PDF.].

#### Getting Started with R Markdown

To get you going with R Markdown, we have prepared an introduction that you can
read
[here](https://github.com/stat-lu/dataviz/raw/main/introduction-to-rmarkdown.pdf).

### Template

We have provided an R Markdown template that we want you to use if you use R
Markdown. Download it to `template.Rmd` in your working directory by calling
the following lines of code.

```{r eval = FALSE}
download.file(
  "https://raw.githubusercontent.com/stat-lu/dataviz/main/template.Rmd",
  "template.Rmd", # filename for the file (which you can replace if you want)
  mode = "wb"
)
```

### WYSIWYG Editors (Office Packages)

If you prefer to use a WYSIWYG editor, please see the Figure Sizing section in
[the text on figure formatting]() for information on how to export and size
your figures appropriately.

#### Pros

* You probably already are familiar with Microsoft Word or Google Docs, so you
  can get going quickly.

#### Cons

* Modifying figures is a little bit of a hassle since you need to re-import the
  figure every time you modify the plot or make changes to its size.
* When you want (or need) to include code, you need to format it yourself by
  using a mono-spaced font (such as Courier or Consolas), and will not get
  syntax highlighting.

## Detailed Requirements

### Submissions

You need to submit a PDF with your work. If you'd like to, however, you are
also free to include `.Rmd` and `.R` files with your submissions containing
code to reproduce your work. Doing so, however, is not at all obligatory.

### Writing

Your report does not need to be verbose but it must be **clear** and
**coherent**. Use full sentences and describe the steps you take and code
you've included. Use a walk-through approach, where you guide the reader
through the report. Someone who has not seen the assignment instructions but
with the same background as you should be able to read and follow along
reasonably well. Do not use a question: answer format in your report.

### Figures

Please see [x]() for detailed information about formatting figures.

### Code

You only need to include code when you are explicitly instructed to do so. When
you do include code, please avoid having it escape into the margin by using
multiple lines to separate code. Good code style is not a requirement in this
course but is nevertheless good practice. It might be a good idea to conform to
a style guide such as [the tidyverse style
guide](https://style.tidyverse.org/). R Studio can also format code for you if
you select it and hit `Cmd/Ctrl + Alt + a`.

## Peer Review and Examination

For each assignment you will need to peer review **three other reports**. To
complete these reviews, you need to fill out the **rubric** associated with the
assignment and add a general comment on the submission. In addition, we also
encourage you to provide **annotated feedback** using the integrated tools in
Canvas. Please see [this
page](https://community.canvaslms.com/t5/Student-Guide/How-do-I-submit-a-peer-review-to-an-assignment/ta-p/293)
to read more about peer-reviewing in canvas.

Finally, please remember to provide **constructive** feedback.

### Grading

The assignments and project will be graded by an instructor or teaching
assistant as **pass** or **fail**. To pass the course, **you need to pass each
assignment**. If you fail any assignment (or the project), you will get an
additional chance to hand in these after the conclusion of the course.

## Hints

- Functions in the tidyverse are well-documented. Students who are successful
  at these assignments often make good use of the help files for the various
  functions. (In R Studio simply place the cursor on the function name and hit
  F1).
- To run code in R Studio inside a R Markdown document, highlight the code and
  hit `Ctrl/Cmd + Enter`. (There is no need to copy it directly into the
  terminal.)
- To create a code chunk in your R Markdown document, use the shortcut `Ctrl/Cmd
  + Alt + i`.
- If you want code in your R Markdown document to appear inside your final
  report, use `echo = TRUE` in the header of the code chunk.
- The size of figures can be controlled using the `fig.width` and `fig.height`
  options in the code header (if you use R Markdown) and the `width` and
  `height` arguments in `ggsave()` if you use a WYSIWYG editor. **Please don't
  resize the images inside the editor**, which will risk making the annotation
  in the visualization hard to read.
