---
title: "One Variable"
subtitle: "Data Visualization"
author: "Johan Larsson"
institute: "The Department of Statistics, Lund University"
date: ""
bibliography: "assets/references.bib"
output:
  xaringan::moon_reader:
    css: [default, "assets/styles.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      titleSlideClass: [middle, center]
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dev = "png",
  fig.width = 5,
  fig.height = 4.3,
  fig.align = "center",
  fig.retina = 3,
  dev.args = list(pointsize = 16),
  crop = TRUE
)

library(tidyverse)
library(tactile)
library(lattice)
library(latticeExtra)
library(kableExtra)

kable <- function(x, ...) {
  knitr::kable(x, escape = FALSE, linesep = "", booktabs = TRUE)
}

library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib("assets/references.bib")

if (!interactive()) {
  theme_set(theme_grey(base_size = 16))
}

xaringanExtra::use_search(show_icon = TRUE)

options(digits = 3)
```

## Visualizing a Single Variable

best place to start: continuous or discrete one-dimensional data

often need **statistical transformation** or **position adjustment**

--

Dotplots are great for small and discrete data.

```{r, fig.height = 2.5, fig.width = 7, fig.cap = "Duration between eruptions of the Old Faithful geyser."}
library(tidyverse)

ggplot(faithful, aes(waiting)) +
  geom_dotplot(binwidth = 1) #<<
```


---

## Histograms

.pull-left[
For larger datasets it is often better to
use a statistical transformation.

Histograms accomplish this by **binning** and **summarizing** (counting).

**caution:** choice of bins may introduce bias
]

.pull-right[
```{r, fig.cap = "Histogram of Old Faithful data."}
faithful %>%
  ggplot(aes(waiting)) +
  geom_histogram(bins = 15) #<<
```

]

---

## Density Plots

attractive for variables that are continuous

but sensitive to settings (type of kernel, bandwidth, and more)

.pull-left[
```{r, fig.height = 3}
faithful %>%
  ggplot(aes(waiting)) +
    # tighter bandwidth
    geom_density(bw = 5) #<<
```
]

.pull-right[
```{r, fig.height = 3}
faithful %>%
  ggplot(aes(waiting)) +
    # longer bandwidth
    geom_density(bw = 10) #<<
```
]

often good idea to add a **rug** layer (`geom_rug()`) to density plots

---

## Boxplots

.pull-left[
most common type:

  - **median** as middle bar, 
  - 1st and 3rd **quartiles** as edges of box,
  - whiskers to last observation within 1.5 times the
    inter-quartile range, and
  - points (or stars) beyond whiskers

useful when visualizing categories or many variables at once

more abstraction than histogram or densityplot 

not suitable for data with multiple modes
]

.pull-right[
```{r, fig.height = 3, fig.cap = "Boxplot of the Old Faithful data, which fails completely in accurately describing the distribution."}
faithful %>%
  ggplot(aes(waiting)) +
    geom_boxplot() #<<
```
]

---

## Violin Plots

a type of density plot, often used instead of boxplot

same pitfalls as with other density plots

```{r, fig.height = 3, fig.width = 6.5, fig.cap = "geom_violin() does not work with a single variable, so we need to use the trick y = 1 here."}
faithful %>%
  ggplot(aes(waiting, y = 1)) +
    geom_violin() #<<
```

---

## Several Variables at Once

boxplots and violin plots are useful for visualizing multiple
variables at once

need to first transform data into **long** format

```{r}
pivot_longer(mpg, c(displ, cty, hwy)) %>% #<<
  ggplot(aes(name, value)) +
    geom_boxplot()
```

---

### Facets

mostly used to visualize **small multiples** by splitting the plot into facets on a 
categorical variable

but can also be used to visualize multiple variables at once

for this use case, it is normally good to set `scales = "free_x"` (or sometimes `"free"`)

```{r fig.width = 9, fig.height = 3}
pivot_longer(mpg, c(displ, cty, hwy)) %>%
  ggplot(aes(value)) +
    geom_histogram(bins = 10) +
    facet_wrap("name", scales = "free_x") #<<
```

