---
title: "Template"
subtitle: "STAE04/DABN19: Data Visualization"
author: "Your Name"
affiliation: "Department of Statistics, Lund University"
date: "`r Sys.Date()`"
papersize: a4
geometry: margin=4cm
colorlinks: true
output:
  pdf_document:
    number_sections: true
---

```{r setup, include = FALSE}
# This is the setup chunk. It can be used to setup general options and load
# packages that you make use of during the report.
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 2.5,
  fig.height = 2.2
)

# set a theme, including font size, for ggplot2
theme_set(theme_grey(base_size = 10))

options(
  digits = 3, # limit number of significant digits
  width = 63  # limit width of code to prevent overflowing
)
```

# Introduction

In Table 1, we outline the properties of some presidents from the USA.

Table: Properties of some presidents in the US and their terms.

Name                   Party         Start of term    End of term
--------------------   -----------   --------------   --------------
Dwight D. Eisenhower   Republican              1953             1961
John F. Kennedy        Democratic              1961             1963
Lyndon B. Johnson      Demoratic               1963             1969
Richard Nixon          Republican              1969             1974

We could also make a neatly formatted table directly from R, using
`knitr::kable()`! Here we explicitly tell R Markdown to show our code
in the output. See the result in Table 2.

```{r r-table, echo = TRUE}
library(tidyverse)

# Pull some data from the presidential data set
presidential %>%
  head(4) %>%
  knitr::kable(
    caption = "The terms of some presidents in the US.",
    booktabs = TRUE # nicer PDF tables (always use this)
  )
```

To wrap up our investigation of US presidents, let's create a plot outlining
the terms of some of the presidents (Figure 1).

```{r plot1, fig.cap = "The terms of some presidents in the US.", fig.width = 5}
presidential %>%
  head(9) %>%
  mutate(order = 1:n(),
         name = reorder(name, order)) %>%
  ggplot(aes(xmin = start, xmax = end, y = name, color = party)) +
  geom_linerange() +
  scale_color_manual(values = c("navy", "red")) +
  labs(y = "President", x = "Time", color = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.pos = c(0.85, 0.25))
```
