---
title: "Worked Example: Heavy Metal"
author: "Johan Larsson"
#date: "9/4/2020"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

## Meuse River Data

In this worked example, we'll take a look at the *Meuse river data set*
from the [sp](https://CRAN.R-project.org/package=sp) package^[The sp package
is a full-featured solution for visualization data, and can be used in its
own right to create maps in R.]

You of course need to download and install the package if you don't have
it already. Then make the dataset available using `data(meuse)`.

```{r}
# install.packages("sp")
library(sp)
data(meuse)

head(meuse)
```

The data consists of longitude (`x`) and latitude (`y`), as well as
different measures related to heavy metals. Before we dig into this
data set, let's set up the ggplot2 theme for mapping. This time, we use
the `theme_map()` (instead of `theme_void()`, which we used
previously) from the
[ggthemes](https://CRAN.R-project.org/package=ggthemes) package.

```{r}
library(tidyverse)
library(ggthemes) # install.packages("ggthemes")

theme_set(theme_map(base_size = 11))
```

Let's start with a simple plot, looking at lead, creating a bubble plot by
mapping the presence of lead to size of the points.

```{r, fig.cap = "Concentration of lead around the river Meuse.", fig.width = 4, fig.height = 5}
ggplot(meuse) +
  aes(x, y, size = lead) +
  geom_point() +
  coord_equal() +
  theme(legend.position = "right")
```

## Raster Charts

Okay, that looks alright, but what we're missing is of course some geographical
features. It's not even clear right now where the river is. To get some
geographical data, we'll download a suitable raster map from stamen maps,
using **ggmap**.

```{r}
library(ggmap) # install.packages("ggmap")
```

To download the map, we of course need to specify coordinates, which
we can find from the `meuse` dataset. The problem is that the
coordinates from the dataset are not standard latitude and longitude
coordinates. Instead, by looking at the documentation for `meuse`, we see
that they are in fact Rijksdriehoek (RDH) (Netherlands topographical) map
coordinates. We need to convert these. Doing so is a little bit
tricky. We first need to setup our original coordinates and define the
coordinate system that they are tied to, then project the coordinates to a
new coordinate system.

This next bit requires the [rgdal](https://CRAN.R-project.org/package=rgdal)
package, which you may need some external dependencies to install.^[Possibly
<https://cran.r-project.org/bin/windows/Rtools/> if you are on Windows and
the development tools if you are on Mac OS X
<https://cran.r-project.org/bin/macosx/tools/>. If you use linux, you should
receive a helpful error message when trying to install the package about
what to do.] If you don't want to or have trouble downloading **rgdal**, we
have provided the converted dataset at the course repository, and
you can load it by calling the following lines.

```{r, eval = FALSE}
meuse2 <- read_csv(
  "https://github.com/stat-lu/dataviz/raw/main/data/meuse2.csv")
```

Otherwise, we now proceed to convert our coordinates.

```{r}
# install.packages("rgdal")
library(rgdal)

rdh <- select(meuse, x, y)
coordinates(rdh) <- ~ x + y
proj4string(rdh) <- CRS("+init=epsg:28992")

lonlat <- spTransform(rdh, CRS("+init=epsg:4326"))
```

Now we have our longitude and latitude coordinates, and simply
need to replace our original coordinates with these.

```{r}
meuse2 <- as_tibble(lonlat) %>%
  bind_cols(select(meuse, -x, -y))
```

Now we extract the boundaries of the data.

```{r}
rng <-
  meuse2 %>%
  summarize(x = extendrange(x), # extendrange() extends the range
            y = extendrange(y))
```

Now we can simply retrieve the map from Stamen maps using **gmap** and
`get_stamenmap()`.

```{r}
mp <- get_stamenmap(
  c(left = rng$x[1],
    right = rng$x[2],
    top = rng$y[2],
    bottom = rng$y[1]),
  zoom = 14,
  maptype = "terrain"
)
```

To then build the map, we use the `ggmap()` function as we would use `ggplot()`,
but make sure to use the `data` argument in our geom (here `geom_point()`)
for our original data of heavy metal concentrations.

```{r, fig.cap = "Concentration of lead around Meuse.", fig.height = 6}
ggmap(mp) +
  geom_point(aes(x, y, size = lead), data = meuse2) +
  theme(legend.position = c(1, 0),
        legend.justification = c("right", "bottom"))
```

## More Heavy Metal

Let's make the plot slightly more interesting by including data on the
other heavy metals too. First, we pivot our dataset to put it in a long
format for ggplot2.

```{r}
meuse3 <-
  meuse2 %>%
  pivot_longer(cadmium:zinc, names_to = "metal", values_to = "concentration")
```

Then we plot as before, faceting on the type of metal.

```{r, fig.cap = "Concentration of different heavy metals around the river Meuse in the Netherlands.", fig.width = 5, fig.height = 6}
ggmap(mp) +
  geom_point(aes(x, y, size = concentration),
             alpha = 0.5,
             data = meuse3) +
  facet_wrap("metal") +
  theme(legend.position = "right")
```

## Geocoding

Purely for instructional reasons, let's assume that we were considering some
kind of cleanup of this heavy metal waste, but wanted to see how this
would affect affect the ruins of the nearby *Kasteel Stein* by
displaying it on the map.

We don't know the coordinates of the castle off-hand, so let's find them out
by geocoding. In the lecture, we used the facilities of **ggmap** for
geocoding, but unfortunately the only API that is currently supported
in **ggmap** is Google's API, which is not free.

Another alternative is to use
[tidygeocoder](https://CRAN.R-project.org/package=tidygeocoder), which
features a host of geocoding alternatives. A couple of these are free, for
instance the [Nominatim](https://nominatim.org/) API.^[The API is free, but
it is also rate-limited at one request per second and not as powerful as
some of the other available APIs. If you have a project where you need more
serious geocoding, you should probably get a key for one of the other APIs.]

Start by installing the **tidygeocoder** package.

```{r, eval = FALSE}
install.packages("tidygeocoder")
```

Then generate a `tibble` (or `data.frame`) with the names
and addresses that you want to geocode.

```{r}
library(tidygeocoder)

addresses <- tribble(
  ~ name,          ~ address,
  "Kasteel Stein", "Kasteel Stein, Netherlands"
)
```

Next, pipe these addresses on to `tidygeocoder::geocode()`.^[There is a
namespace clash here with **ggmap**, which has its own `ggmap::geocode()`
function, so we need to specify which function to use with `::` here.]

```{r}
kasteel_stein <-
  addresses %>%
  tidygeocoder::geocode(address,        # map to address column
                        method = "osm") # for Nominatim API
kasteel_stein
```

To finish our map, we complement our previous plot with two layers
for our new, geocoded data, to show a label and a dot for the castle ruins.

```{r, fig.cap = "Concentration of different heavy metals around the river Meuse in the Netherlands.", fig.width = 6, fig.height = 7}
ggmap(mp) +
  geom_point(aes(x, y, size = concentration),
             alpha = 0.5,
             data = meuse3) +
  geom_point(aes(long, lat), col = "red", data = kasteel_stein) +
  geom_label(aes(long, lat, label = name), data = kasteel_stein, vjust = -0.5) +
  facet_wrap("metal") +
  theme(legend.position = "right")
```

## Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-chile-plebiscite.Rmd>.
