---
title: "Worked Example: Hello, World!"
author: "Johan Larsson"
#date: "9/4/2020"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

## Overview

Let's get started working with R! In this worked example, we will learn how
to install and use packages in R as well as perform some basic operations on
vectors, matrices, and data frames. First off, if you're familiar with
other programming languages since before, you will notice that
R is an *interpreted* language, which means that you don't need to compile
anything.^[R actually does a bit of just-in-time compiling for you, but you
will hardly notice.]

That means it's very easy to write and run an R program. You simply put some R
code in a text file with the ending `.R`, and then you can run that program
using a terminal or a graphical user interface like R Studio. So,
let's say we have a script called [`helloworld.R`](helloworld.R). Then it's just
a matter of calling `Rscript`^[`Rscript` is a helper application that is
installed alongside R to make it easy to run scripts.] on that scripts, like
this.

```{bash}
Rscript helloworld.R
```

This is just to showcase that R is really just a program that reads and
interprets text files. The `.R` (and `.Rmd`) files that you will be
working with are just that: text.

## Objects in R

When you start an R session, you enter into an **environment** that stores
objects that you access via their **names**. You can populate this environment
with objects of your own using the `<-` operator by placing
the name (or variable) you want to give your object on the left-hand side, and
the object itself on the right-hand side, like this:

```{r}
# store the number 42 in the_meaning_of_life
the_meaning_of_life <- 42
```

Then to access this object, you simply have to refer to it as
`the_meaning_of_life` (no quotation marks).

You can (usually) print the contents of any variable in R by simply calling its
name in the console.

```{r}
the_meaning_of_life
```

### Vectors

Vectors are one-dimensional arrays which store objects. Vectors can be created
from functions or directly using, for instance, the `c()` function. Each vector
can only store objects of *a single type*, which means that you cannot combine
the different data types in R, such as **character**, **numeric**, and
**integer**. If you still decide to try to, however, you will soon notice that R
won't tell you no. Instead, R simply *converts* the vector to a single type
based on a few defaults.

We talked about this in the lecture, so try to see if you can figure out how the
following vectors will be converted (if at all):

```{r eval = FALSE}
c("a", 1, TRUE)
c(1L, 1) # 1L = integer of value 1
c(NA, "a")
c(1, 1/3)
```

Having numeric values be converted to characters is seldom what you want. The
conversion from integers to doubles if often not a big problem, but can lead
to improper formatting of things such as axis labels in a plot.

We can manipulate vectors algebraically using standard arithmetic operators.

```{r}
1:2 + c(3, 9)      # adding two vectors together
1.6 / 2
c(1, 10) * 2       # 2 is multiplied with each element
c(NA, 3.123, pi)^2 # pi is an object that's always available
```

If you combine two vectors thare are of different length but where the
shorter vector's length is a *multiple* of the longer vector's length, then
R will *recycle* the shorter vector.

```{r}
c(1, 2, 3, 4) * c(1, 2)
```

What do you think happens if the shorter vector's length is *not* a multiple
of the longer vector's length? We'll find out a bit later.

### Special Values

`NA` is a special value used to denote missing data. There is also another type
of value called `NULL`, which is exactly what is sounds like: nothing.

We can assign `NULL` to a variable

```{r}
a <- NULL
```

but note that creating a vector with several values, some of them `NULL`, will
give a different result.

```{r}
a <- c(NULL, "a")
a
```

To access an element in a vector, we can use the `[` operator:

```{r}
x <- c("a", "b", "c", "e")
x[1]
```

And to access several elements at once, you can use a vector of indices.

```{r}
x[c(1, 4)]
```

Accessing a contiguous sub-vector is often desirable, in which case the `:`
operator is a useful tool. It creates a vector of indices.

```{r}
1:4
```

We can use this to access a range of elements from `x`.

```{r}
x[1:3]
```

It is also possible to modify elements in vectors in the same fashion.

```{r}
x[1:2] <- c("foo", "bar")
x
```

Similarly, we can *remove* certain elements by negative indexing in the
following way.

```{r}
x[-1]
x[-(1:2)]
```

Notice the use of parentheses in the second example. What do you think
`x[-1:2]` would give us?

### Matrices

Matrices are two-dimensional arrays. Just like vectors, they can only contain a
single type of values.

```{r}
m <- rbind(c(1, 2),
           c(3, NA))
m
```

We can access separate rows and columns by separating row and column indices
with a comma.

```{r}
m[1, 2]
```

To select an entire column, simply omit the first argument (or vice verse to
select an entire row).

```{r}
m[1, ]
m[, 2]
```

### Data Frames

Data frames look a lot like matrices except they allow each column to have a
different type. Most of the data that you will encounter in this course will be
stored as data frames (or as tibbles, but more on that later).

The following code creates a simple data.frame, using the `data.frame()`
function.

```{r}
d <- data.frame(a = 1:3,
                b = letters[1:3])
d
```

In addition to sub-setting columns using the matrix notation (i.e. `d[, 2]` for
the second row), we can subset columns using `$` as follows.

```{r}
d$a
d$b
```

## Packages

### Installing Packages

One of R's strengths is the wide-ranging package ecosystem it sports.
One package that we will use a lot during this course is **ggplot2**. If you
don't yet have that package installed, we will now do so by calling the
`install.packages()` function, using **ggplot2** as the first argument. Simply
call the code below.

```{r eval = FALSE}
install.packages("ggplot2")
```

When you call this line, R will also install all of the *dependencies* that
come with ggplot2, which includes other packages. If this is the first time you
run R, you might also be prompted to choose a CRAN mirror. A good default choice
is `"https://cloud.r-project.org"`.

### Using Packages

To use a package in R, you can call the `library(<package-name>)` function,
after which all of that packages functions---and often also its
datasets---will be available to you directly in your working environment.

Load the **ggplot2** package we just installed by calling the following.

```{r}
library(ggplot2) # library("ggplot2") would work as well!
```

Sometimes we will also use `<package>::function()` to call a
function from a specific package. In this case, you don't need to call library
first. This is particularly useful if you load two packages that export
two functions with the same names.

In the following line, we create a plot in **ggplot2** with this syntax, but
bear in mind that this is just for show since we're already loaded the package
using `library()`.

```{r}
ggplot2::ggplot(mpg, ggplot2::aes(cty, hwy)) +
  ggplot2::geom_point()
```

## Functions

### Function Basics

As we said in the lecture, you will use functions *a lot* throughout this
course (although you won't write your own). Therefore, you really need to
understand how to *call* functions.

Here is a simple function:

```{r}
f1 <- function(a, b) {
  (a + b) ^ 2
}
```

Its arguments are `a` and `b` and it returns the squared sum of `a` and `b`.
When we call it we must supply both arguments or R will throw an error.

```{r eval = FALSE}
f1(2) # error!
```

```{r}
f1(2, 3)
```

You can choose to specify arguments by location, as we did above, or by
naming each argument as in the following call.

```{r}
f1(a = 2, b = 3)
```

When you use functions in R, however, you will find that you seldom need to
supply all the arguments of a function, which is due to them having
*default values*.

The following function, `f2()`, have defaults for both of its arguments.

```{r}
f2 <- function(a = 3, b = 0.5) {
  sqrt(a) + 3 * log(b)
}
```

We can call it without supplying *any* argument.

```{r}
f2()
```

We could also supply only `a`, by simply using one argument.

```{r eval = FALSE}
f2(0) # error
```

To only supply `b`, you have to name the argument.

```{r}
f2(b = 1)
```

### Vectorization

Many functions in R are **vectorized**, which means that you can apply the
function across an entire vector (or column of a matrix) without
any extra work. All of the arithmetics functions you saw previously, as well
as `log()`, `sqrt()`, and many more, share this feature.

For instance, say that you have the following data frame,

```{r}
d <- data.frame(
  name = c("Wahijd", "Mary", "Eyvind"),
  diet = c("vegetarian", "meat-eater", "vegetarian"),
  age = c(34, 89, 50)
)
d
```

and want to convert the age of the three persons to days instead. We can then
simply modify the entire vector in place by calling

```{r}
d$age <- d$age * 365
```

## Errors and Warnings

### Errors

We have already shown a few lines of code that when run produces **errors**.
Errors notify you that something has gone wrong, and this always happens when
calling a function (remember that everything you do in R is the result of a
function call; even `2 + 3` calls a function under the hood).

With every error, you will receive a message indicating what has gone wrong
(and sometimes why). Error messages can, however, be rather cryptic. The best
advice, if you don't understand what's going on, is to simply search for the
error message ad verbatim in your favorite search engine. Chances are that
someone else has been there before.

### Warnings

You may also occasionally receive *warnings*. Warnings indicate that something
may not have gone as planned but that the problem hasn't been fatal enough to
cause the function to terminate. Be weary of warnings. Warnings can be harmless
but they may also indicate that something isn't working properly, and you will
do well to investigate if that is the case.

Here is a line that throws a warning. Before you run the line, can you figure
out what might cause it to give off a warning? What do you even expect to happen
when running the code?

```{r eval = FALSE}
c(3, 2, 1) * c(1, 2)
```

## Conditional Statements

### Basic Conditional Operators

Conditional statements are very useful when it comes to data science,
particularly for filtering observations based on the values of another variable.
Conditionals are logical statements that check each value to see if it is equal
to, greater/lesser than, or greater/lesser or equal to some reference.

Conditional statements return `TRUE` or `FALSE` and are typically invoked
through the following conditional operators:

operator  meaning
--------  -------------------------
`==`      equal to
`!=`      not equal to
`>`       greater than
`<`       lesser than
`>=`      greater than or equal to
`<=`      lesser than or equal to

Let's see how it works in practice. Try
to predict the output of the following conditional statements before running
them.

```{r eval = FALSE}
3 > 2

1/3 == 0.33 # is this surprising? why not?

TRUE != TRUE

FALSE == FALSE

TRUE == 1

FALSE != 0

"b" > "a"

"ab" < "ac"
```

These statements are also **vectorized**, which means that we can
do the following.

```{r}
c("a", "b", "a") == "a"

1:5 != 3
```

### The `%in%` Operator

Another, very useful, conditional operator is `%in%`. It checks if the left-hand
side of the operator is inside the right-hand side. Observe the following
example.

```{r}
c("a", "b", "c") %in% "a"
```

This is very useful for checking which observations on a variable belong to
a set of levels. Here we re-use the `d` dataset from above to check which
persons are vegans or vegetarians.

```{r}
d$diet %in% c("vegan", "vegetarian")
```

This will come in handy for us repeatedly throughout the course.

## Documentation

The easiest way to understand how a function works, what it returns,
what arguments it has and what they mean, as well as what the default argument
values are is to read the documentation of the function.

If you're using R Studio (which you really should) you can simply place your
cursor on the function, click, and hit `F1` to bring up the documentation for
the function in the help pane of R Studio. It is also possible to use
`help(<function-name>)` or `?<function-name>`.

```{r}
help(mean) # or ?mean
```

## Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-hello-world.Rmd>.
