---
title: "Worked Example: Gapminder"
author: "Johan Larsson"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

## Gapminder

Do you remember the animated plots we produced in the introductory
lecture for this course based on the Gapminder Hans Rosling
animated visualization?

<iframe width="560" height="315" src="https://youtu.be/jbkSRLYSojo?rel=0" frameborder="0" allowfullscreen></iframe>

In this worked example, we'll work out how to
reproduce that plot as both an animated an interactive visualization.

## Animated Visualization

The dataset that we'll use is available via the
[gapminder](https://CRAN.R-project.org/package=gapminder) package. So go
ahead and install that package.

```{r eval = FALSE}
install.packages("gapminder")
```

This makes the dataset available in an object called `gapminder`. These
are the first few rows of the dataset.

```{r}
library(gapminder)

head(gapminder)
```

The variables should be self-explanatory.

Let's jump right in and create a bubble plot faceted on year (which we cut
into groups), with population mapped to the size of the bubbles, and
GDP per capita and life expectancy on the x and y axes respectively.

```{r fig.cap = "Life Expectancy and GDP per capita from 1950 to 2010.", fig.width = 8, fig.height = 7}
library(tidyverse)

gapminder %>%
  mutate(years = cut_interval(year, length = 5)) %>%
  ggplot(aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap("years") +
  labs(y = "Life Expectancy",
       x = "GDP per Capita",
       size = "population")
```

Just as we said in the first lecture, this visualization is not (yet) working
out so well for us. Let's make it animated instead. For this, we'll use
the [gganimate](https://CRAN.R-project.org/package=gganimate) package.
First install the package.

```{r, eval = FALSE}
install.packages("gganimate")
```

To use the **gganimate** package you also need a *renderer* to produce
animated images. You can use either [gifski](https://gif.ski/) or
[ImageMagick](https://imagemagick.org/). We recommend the former (and
**gganimate** defaults to gifski if it is installed), but either
will work just fine. Run one (or both) of the following lines to
install a renderer.

```{r, eval = FALSE}
install.packages("gifski")
install.packages("magick")
```

We build the plot as before, but now make it animated by adding
the `transition_time()` function to the plot as well as use the
title label to show the current year.

```{r, fig.cap = "GDP per capita and life expectancy for some of the countries of the world.", fig.width = 7, fig.height = 6}
library(gganimate)

ggplot(gapminder,  aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  labs(title = "Year: {frame_time}", # special glue syntax
       y = "Life Expectancy",
       x = "GDP per Capita",
       size = "population") +
  transition_time(year)
```

If you think the plot is still crowded, we could alternatively use
facets to separate continents. Here we also make use of the
`country_colors` object that is included in the **gapminder** package.

```{r, fig.cap = "GDP per capita and life expectancy; now with facets!", fig.width = 7, fig.height = 6}
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(alpha = 0.5) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(title = "Year: {frame_time}",
       x = "GDP per capita",
       y = "Life expectancy") +
  transition_time(year)
```

So far our plot does a good job of showing the trends among the various
continents of the world but is hard to use if we are interested in one
specific country. A remedy for this can be to use labels to let us
identify which bubble belongs to which country. The large number of countries,
however, means that it's not a frightfully good idea to label *all* of them.

Instead, we'll pick out the largest two countries (at the latest
time stamp) on each continent and label those. First, we
store the names of the countries in a vector, `large_country_names`.

The following steps first filter the dataset so that only
observations from the latest year (`max(year)`) are kept, then groups the
dataset by continent, then slices the dataset so that the observations
(countries) with the largest and next-to-largest values of population (`pop`)
of each group (continent) are kept, and then finally pulls out (using `pull()`)
the country names.

```{r}
large_country_names <-
  gapminder %>%
  filter(year == max(year)) %>%
  group_by(continent) %>%
  slice_max(pop, n = 2) %>%
  pull(country)

large_country_names
```

Then we filter the original dataset to create a separate dataset for our
labels.

```{r}
large_countries <-
  filter(gapminder, country %in% large_country_names)
```

Now we put everything together; this time we also change the
*easing* of aesthetics from linear to cubic in-and-out using `ease_aes()`,
to more clearly show that we actually only have data on a 5-year interval here.
We label the countries with `geom_label_repel()` from the **ggrepel** package,
in order to avoid overlapping labels.^[Working with labels and animated
visualizations is something of a challenge. Here I've had to tweak the
settings (mostly `nudge_x` and `nudge_y`) several times in order to get
something that looks good.]

```{r, fig.cap = "Life expectancy and GDP per capita with countries. The two largest countries at the start (in terms of population) have been labeled.", fig.width = 8, fig.height = 7}
library(ggrepel)

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(alpha = 0.5) +
  geom_label_repel(
    aes(gdpPercap, lifeExp, label = country),
    inherit.aes = FALSE,
    seed = 1, # important when animating
    nudge_x = 5,
    nudge_y = -10,
    data = large_countries
  ) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(title = "Year: {frame_time}",
       x = "GDP per capita",
       y = "Life expectancy") +
  transition_time(year) +
  ease_aes("cubic-in-out")
```

## Interactive Visualization

Interactive visualizations are often effective, particularly when
we want to visualize a complicated dataset such as this one.
Here we'll use the [plotly](https://CRAN.R-project.org/package=plotly) package
to do so, which, as you may recall from the lecture, works well in tandem
with ggplot. First install the package.

```{r, eval = FALSE}
install.packages("plotly")
```

Then load the package.

```{r}
library(plotly)
```

Now we redraw the plot, adding an interactive slider to select the year using
plotly. Make note of the additional mapping that we've
added to `geom_point()`, namely `frame`, which is a special
mapping that will let plotly know which variable to use to separate
the visualization into frames.

```{r, fig.cap = "An interactive visualization using plotly for the Gapminder data.", fig.width = 8, fig.height = 7}
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(aes(frame = year), alpha = 0.5) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(x = "GDP per capita",
       y = "Life expectancy")

ggplotly(p)
```

Notice how seamless the conversion of ggplots into interactive plots can be
with the help of plotly.

## Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-gapminder.Rmd>.
