---
title: "Worked Example: First Steps with ggplot2"
author: "Johan Larsson"
#date: "9/4/2020"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

## ggplot2

ggplot2 is part of the **tidyverse** so you can simply call

```{r}
library(tidyverse)
```

to load ggplot2.

## `ggplot()`

The main function of ggplot is the `ggplot()` function. It has two main
arguments that you need to be aware of:

`data`
: the (default) dataset to use for the plot

`mapping`
: the aesthetic mappings

When calling `ggplot()` we typically don't call these arguments by their name
and instead simply supply them in order.

### `data`

The `data` argument obviously needs a dataset and this should usually come in
the form of a `data.frame` or a `tibble`. Let's starts with a very simple data
set, `cars`, that contains only two variables: speed and stopping distance of 50
different cars.

Before doing anything else, always take a little time to look at bit of the
data. If the dataset has relatively few variables (less than ten), the `head()`
function is usually your best option; it simply displays the first few rows of
the data.

```{r}
head(cars)
```

The second argument of `heads()` decides how many rows to show. Modify this
value to show the first 10 cars of the dataset.

`glimpse()` is very useful when you have more variables (since it is more
compact). Call `glimpse()` on `cars` and notice that it lists the variables in a
table along with the first values of each variable.

### `mapping`

The `mapping` argument should be supplied with a call to the `aes()` function
that lists the mappings between variables in your data and aesthetics of the
plot. Which aesthetics you choose to map to depends on what geoms you intend to
use. But for now, let's assume that we want to map speed (`speed`) to the x
axis, and stopping distance (`dist`) to the y axis. To do so, we simply use
`aes(x = speed, y = dist)` as the input to the `mapping` argument. Because the
first two arguments of `aes()` are `x` and `y`, we can do without calling them
explicitly.

Let's put this together:

```{r blank-canvas}
ggplot(cars, aes(speed, dist))
```

Why does this result in a blank canvas? After all, we did tell ggplot to map
speed to the x axis and stopping distance to the y axis?

The answer is that we still haven't told ggplot2 which layers we want to use.

## Layers

The aesthetic mapping is not enough by itself---we also need a layer to display
our data. In **ggplot2**, we accomplish this by adding (stacking) layers to the
output of `ggplot()`. There are several ways to construct layers in ggplot, but
the simplest and most common approach is to use the functions named `geom_*()`
(such as `geom_point()`).

The names of these `geom_*()` functions are actually misnomers, given that the
they in fact set up entire layers and not just the geoms.

To provide a layer with the point geom to the plot, you simply need to call
the following lines.

```{r cars-plot}
ggplot(cars, aes(speed, dist)) +
  geom_point()
```

Notice the use of `+` here to add layers to the canvas that we started out with.
Also observe that we didn't need to specify what data to use or which mappings
we wanted in the call to `geom_point()`---they were inherited from the
`ggplot()` call. If you want to avoid this type of inheritance (which you
will want when constructing plots from different data sources), you can
do so by specifying `inherit.aes = FALSE` in the call to `geom_point()`.

Try to switch geom for your plot. Replace `geom_point()` by `geom_line()`.

Now let's try using `geom_rect()` instead.

```{r cars-plot-fail, eval = FALSE}
ggplot(cars, aes(speed, dist)) +
  geom_rect()
```

Well that didn't work, but why not? Well, the reason is simply that
`geom_rect()` uses the rectangle geom and a rectangle needs more aesthetics than
just x and y coordinates.

## Aesthetics

Look at the documentation of the `geom_point()` function. Under **Aesthetics**
we can see that this layer understands the following aesthetics:

- **x**
- **y**
- alpha
- colour
- fill
- group
- shape
- size
- stroke

and that only the two first are in fact required. The remaining aesthetics have
been supplied defaults, such as `alpha = 1` (no transparency) and
`colour = "black"` (black points). But these aesthetics can also be changed,
either by mapping them to another variable in our data or by manually
adjusting them directly.

When you change an aesthetic by specifying a value (and not a mapping) you
simply specify that aesthetic in the call to the geom you would like to change.
Let's change the color of the points to navy blue, increase the size, and change
shape. Doing so is a breeze with ggplot2:

```{r cars-navy}
ggplot(cars, aes(speed, dist)) +
  geom_point(colour = "navy",
             size = 1.8,
             shape = 4)
```

Take some time to play around with the various aesthetics. Switch geom and try
to flip the axes. Can you figure out what the `stroke` aesthetic does?

It might be worth noting that some of the aesthetics used here have synonyms
that you will likely come across every now and then, for instance `cex = size`,
`col = color = colour`, and `pch = shape`, which are there mostly for
reasons that are historical or have to do with compatibility.

## Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-first-steps-with-ggplot2.Rmd>.
