---
title: "Worked Example: Arthritis"
author: "Johan Larsson"
#date: "9/4/2020"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

## Arthritis

In this worked example we will be working with data from a treatment trial for
Arthritis. The data comes from the [vcd
package](https://CRAN.R-project.org/package=vcd). To begin, let's load the
package and take a `glimpse()` at the data.

```{r}
# install.packages("vcd") # if you haven't installed it yet
library(vcd)
library(tidyverse)

glimpse(Arthritis)
```

Notice that `Improved` is helpfully formatted as an ordinal (ordered) variable
already, which is going to help ggplot to choose the correct color palettes for
us.

## Bar Charts

Let's start with a simple bar chart to see how many people were allocated to the
two treatment arms.

```{r, fig.cap = "Treatment allocation for the arthritis study.", fig.width = 2.2, fig.height = 2.2}
library(tidyverse)

ggplot(Arthritis, aes(Treatment)) +
  geom_bar()
```

### Was the Treatment Effective?

Now for something more interesting. Let's try to look at how well the treatment
seems to have worked. We start with a stacked bar chart.

```{r, fig.cap = "Treatment effect in the arthritis study.", fig.height = 3}
ggplot(Arthritis, aes(Treatment, fill = Improved)) +
  geom_bar() +
  labs(fill = "Improvement") # add a better legend title
```

It's immediately clear that the treatment has worked well! The count of people
with a improvement is much higher in the treatment arm compared with the placebo
arm. But how do the treatments compare when it comes to people who've made
*some* improvement? This is difficult to say from our approach here. This
is due to two reasons:

1. The stacked bar chart makes it very hard to compare all but the
   first category, since the other portions of the bars are not aligned.^[This
   is not quite as severe a problem for this particular visualization since the
   inner grouping variable is ordinal. For other problems, however, this can
   be much more detrimental to the result.]
2. The bars are not proportional.^[To be fair, the differences between the group
   sizes are small here, but this also means that we don't lose much information
   when using proportional bars instead.]

Let's try a grouped proportional bar chart instead. To do this, we need
to do a little preparatory work since there is no direct option to do this
in ggplot2. 

What we need to do is to prepare a data set with the proportions that we would
like to plot. In the following lines, we

1. use `count()` to count the number of observations in each combination of
   `Treatment` and `Improved`,
2. group the data set by `Treatment`, and
3. compute the proportions using `mutate()`.

```{r}
arthritis_prop <-
  Arthritis %>%
  count(Treatment, Improved) %>% # like group_by + summarize
  group_by(Treatment) %>%
  mutate(prop = n / sum(n))
```

Now we're ready to construct our plot, which we do exactly like a standard
grouped bar plot, except we use `geom_col()` and map the `y` aesthetic since we
have summary data.

```{r, fig.cap = "Grouped, proportional bar chart of the arthritis data.", fig.height = 3}
ggplot(arthritis_prop, aes(Treatment, prop, fill = Improved)) +
  geom_col(position = "dodge") +
  labs(fill = "Improvement", y = "Proportion")
```

No it's suddenly quite easy to compare the proportion of people who improved
some between the two groups. 

## Mosaic Plots

In the lecture we also encountered mosaic plots, which are proportional bar
charts that map not only the height of the bars but also the width of the bars
to group sizes.

In the lecture, we used the
[productplots](https://CRAN.R-project.org/package=productplots) package, but
here we will try another solution offered by
[ggmosaic](https://CRAN.R-project.org/package=ggmosaic).

```{r}
# install.packages("ggmosaic") # if don't already have the package

library(ggmosaic)
```

Unlike **productplots**, **ggmosaic** works directly with a ggplot2 object,
adding auxiliary functions (`geom_*()` functions) to extend its functionality.

```{r}
ggplot(Arthritis) +
  geom_mosaic(aes(x = product(Treatment), fill = Improved))
```

The key features to notice about this call are

1. that the `aes()` specification must reside *inside* the call to
   `geom_mosaic()`, and not in the `ggplot()` function and
2. the call to `product()`, which acts simply as a list of variables that
   you want to use in the mosaic plot.

The main benefit of the mosaic plot is that we get to see more of the data. With
the bar charts we needed to separate visualizations to provide information on
the overall size of the groups (treatment arms), whereas with the mosaic plot we
can make do with only a single plot.^[But note that it is the *relative* size
that we show. If we want the actual *counts* in the outer category, we still
need to add additional plots or provide this information in text.]

## Waffle Plots

Waffle plots are easy-to-read visualizations that are actually quite uncommon
in statistical visualizations but much more prevalent in media outlets such as
newspapers. They are, as we mentioned in the lecture, at their best when some of
the categories are so small that their sizes are hard to estimate in a
bar chart or mosaic plot.

There is a great package for mosaic plots that interact well with ggplot2 called
[waffle](https://CRAN.R-project.org/package=waffle). Unfortunately, the current
version on CRAN is outdated^[Feel free to voice your support for a CRAN update
at <https://github.com/hrbrmstr/waffle/issues/72>.] and much of the new
functionality that makes it work well with ggplot is only available from the
[development version at GitHub](https://github.com/hrbrmstr/waffle).

For that reason, waffle plots are not compulsory material in this course and
will instead serve as bonus material for those who are interested.

To download development versions of R packages from GitHub, you need the 
[remotes](https://CRAN.R-project.org/package=remotes) package or the
[devtools](https://CRAN.R-project.org/package=devtools) package.

```{r, eval = FALSE}
# install remotes
install.packages("remotes")

# install waffle
remotes::install_github("hrbrmstr/waffle") # or devtools::install_github()
```

The waffle package requires that our data contains counts of the categories, so
before we can use it, we need to do a bit of wrangling. More precisely, you need
to group the data by `Improved` and `Treatment` and then use `summarize()`
together with either the `n()` function to get either counts or some other
expression for proportions. We'll settle for counts in this example. Before
looking at the following code, see if you can do this by yourself. The result
should be a 6-row tibble with columns `Treatment`, `Improved` and `Count` (or
whatever you want to call it). Save the result to an object called
`arthritis_summarized`.

```{r}
arthritis_summarized <-
  Arthritis %>%
  group_by(Treatment, Improved) %>%
  summarize(Count = n())
```

Now we're ready to start making waffles! The aesthetics you need to know about
here are `fill` and `values`. Values is what we want to map the counts or
proportions that we just created to and `fill` will connect our inner-most group
to fill color. We start with the simplest possible waffle chart, just looking at
improvement among the participants in the treatment arm.

```{r, fig.cap = "A basic waffle chart of the results from the treatment arm."}
library(waffle)

filter(arthritis_summarized, Treatment == "Treated") %>%
  ggplot(aes(fill = Improved, values = Count)) +
  geom_waffle() +
  labs(fill = "Improvement")
```

This looks kind of like a waffle chart, but there's a lot of superfluous
information that we'd rather get rid off. Thankfully, **waffle** provides the
helpful `theme_enhance_waffle()` function for this. Moreover, we probably prefer
the symbols (rectangles) to be squares---it's supposed to look like a waffle
after all, right?---which we can accomplish by forcing the coordinate system to
be isometric using `coord_equal()`. Finally, I think it looks more aesthetically
pleasing if we swap the black line color for white, increase their widths
slightly, and change theme to the minimal variant.

```{r, fig.cap = "A prettier waffle plot."}
filter(arthritis_summarized, Treatment == "Treated") %>%
  ggplot(aes(fill = Improved, values = Count)) +
  geom_waffle(color = "white", size = 1.5) +
  coord_equal() +
  theme_minimal() +
  theme_enhance_waffle() +
  labs(fill = "Improvement")
```

That looks much better, doesn't it?

To wrap up, we now look to the full data set by including both treatment arms.
To do this with **waffle**, you need to use facets. Here we do so via the
`facet_wrap()` function, simple providing the outer category that we want to
facet on.

```{r, fig.cap = "Improvement after treatment for arthritis in placebo and intervention arms.", fig.width = 6, fig.height = 2}
ggplot(arthritis_summarized, aes(fill = Improved, values = Count)) +
  geom_waffle(color = "white", size = 1.5, n_rows = 5) +
  facet_wrap("Treatment") +
  coord_equal() +
  theme_minimal() +
  theme_enhance_waffle() +
  labs(fill = "Improvement")
```

If you want to make other types of pictograms (with other glyphs instead
of squares), be sure to check out the examples over at
<https://github.com/hrbrmstr/waffle>.

## Euler Diagrams

Euler diagrams are used to visualize relationships between sets. Most of
you have probably encountered Venn diagrams during in your first
statistics courses. Euler diagrams are like Venn diagrams but don't have the
requirement that we need to visualize intersections that are empty. Euler
diagrams are only really useful when you have overlap among categories, such
that observations can belong to any, all, or none of the levels of a category.

Here we'll use the [eulerr](https://CRAN.R-project.org/package=eulerr) package.
This package creates *proportional* Euler diagrams, trying to make the
area of each intersection proportional to that intersection's size. This is
not always possible, in which case `euler()` will return an approximate fit.
For this reason, it is important to always first *fit* the diagram, inspect
the fit, and then finally plot it.

The simplest way to run `euler()` is to provide a named numeric vector
that indicates intersections and their sizes. Here's a simple example.

```{r}
# install.packages("eulerr")
library(eulerr)

fit <- euler(c("A" = 5, "B" = 2, "A&B" = 1))
```

Notice the use of the ampersand (`&`) to separate intersections here. Let's
examine the fit of the Euler diagram.

```{r}
fit
```

Stress and diagError are both 0, so the fit is perfect! This will in fact
always be the case for two-set combinations. No we can safely plot the result.
Simply call `plot()` on `fit` to do so.

```{r, fig.cap = "A simple Euler diagram.", fig.width = 3, fig.height = 3}
plot(fit)
```

`euler()` can also work directly with data, provided that the columns are
factors, characters, or logicals. The arthritis data that we've dealt with
so far, however, does not lend itself well to an Euler diagram. Can you
think of why?

The reason is that most of our categories are exclusive. A person cannot be
both in the treatment arm and in the placebo arm, or for that matter
have made both no improvement and a marked improvement, so it is not
suitable for this kind of visualization.

Instead, we'll take a look at a data set from the **eulerr** package
called `organisms`. It contains features from a list of organisms:

```{r}
organisms
```

Each value indicates for a given organism, such as a cat, whether it is
an animal (yes), a mammal (yes), plant (no), and so on. We can simply
call `euler()` on this data set directly.

```{r}
(fit_organisms <- euler(organisms)) # parentheses makes R print the object
```

This doesn't look great, unfortunately. The `stress` and `diagError`
are large^[We won't delve into what these actually mean here, but please read
the documentation for the package if you are interested.] and if we
take a look at a diagnostic plot, we'll see that many of the areas in the
plot are misrepresented.

```{r, fig.cap = "A diagonstic plot of the fit on the organisms data."}
error_plot(fit_organisms)
```

One alternative we can try is to use *ellipses* instead of circles. We
specify this in the call to `euler()`, like so:

```{r}
(fit_organisms_ellipses <- euler(organisms, shape = "ellipse"))
```

The stress metric has improved significantly, but not the diagError measure.
Looking at the error plot, it seems like the errors are smaller than before.

```{r}
error_plot(fit_organisms_ellipses)
```

At this point, we need to think hard about whether we can tolerate this
type of error or not---notice that there, for instance, still is overlap between
the plant and mammal categories. If not, we should look for another type of
visualization altogether, but if yes, then we are now ready to plot the final
result.

```{r, fig.cap = "Relationships between various categorizations of organisms."}
# quantities = TRUE shows the number of actual observations in each
# intersection
plot(fit_organisms_ellipses, quantities = TRUE)
```

## Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-arthritis.Rmd>.

