---
title: "Worked Example: Chile Plebiscite"
author: "Johan Larsson"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

## The 1988 Chilean Referendum

The [1988 Chilean national plebiscite](https://en.wikipedia.org/wiki/1988_Chilean_national_plebiscite)
ended the reign of Chile's then president (and former dictator) August Pinochet.
Those in opposition to extend Pinochet's rule for another eight
years won with 56% of the vote. In this worked example we'll try to probe
the results of this plebiscite (referendum) by looking at a survey of voting
intentions conducted a couple of months prior to the election.

The dataset is available in the
[carData](https://CRAN.R-project.org/package=carData) package. Start by
installing it (if you haven't already).

```{r, eval = FALSE}
# install the carData package
install.packages("carData")
```

Then load the dataset directly into your workspace by calling the
`data()` function as shown next.

```{r}
data("Chile", package = "carData")
```

There are eight variables in this dataset:

`region`
: A factor with levels: C, Central; M, Metropolitan Santiago area; N, North; S,
  South; SA, city of Santiago.

`population`
: population size of respondent's community

`sex`
: a factor with levels: F, female; M, male

`age`
: age (in years)

`education`
: a factor with levels (note: out of order): P, Primary; PS, Post-secondary;
  S, Secondary

`income`
: monthly income (in Pesos)

`statusquo`
: scale of support for the status quo

`vote`
: a factor with levels: A, will abstain; N, will vote no (against Pinochet);
  U, undecided; Y, will vote yes (for Pinochet).

## Wrangling

Before we start visualizing this dataset, we need to format it properly.
The names of the variables are fine, but the factor levels are hard to
decipher. The `education` variable is also not coded as an ordinal variable
although it should be. We save the new dataset in a new object called `Chile2`.

```{r}
library(tidyverse)

Chile2 <-
  Chile %>%
  as_tibble() %>%
  mutate(
    region = recode(
      region,
      "C" = "Central",
      "M" = "Metropolitan Santiago",
      "N" = "North",
      "S" = "South",
      "SA" = "City of Santiago"
    ),
    sex = recode(sex, "F" = "Women", "M" = "Men"),
    education = recode_factor(
      education,
      "P" = "Primary",
      "S" = "Secondary",
      "PS" = "Post-Secondary",
      .ordered = TRUE
    ),
    vote = recode(
      vote,
      "A" = "Abstain",
      "N" = "No",
      "U" = "Undecided",
      "Y" = "Yes"
    )
  )

Chile2
```

That looks much better.

Now let's start by simply checking what the people in this survey actually
intended to vote.

```{r}
ggplot(Chile2, aes(vote)) +
  geom_bar()
```

As we can see the referendum seem to be engaging much of the population under
study. The no voters lead by a slim margin here, but there are many who are
yet undecided.

## Age and Voting Intentions

Let's try to see if we can flesh out these differences more in detail. We
begin by studying age and voting intentions. Our first attempt is
a proportional stacked density plot here, to
focus on the proportions of voting intentions within each age group.

But before we do so, we will switch to a better default color palette for
categorical (discrete) data, since the default palette in ggplot2 is
really not quite ideal. A good choice could for instance
be the "Accent" palette from
[RColorBrewer](https://CRAN.R-project.org/package=RColorBrewer),
but to spice it up we'll go with a color palette inspired from
The Darjeeling Limited by Wes Anderson that is available
via the [wesanderson](https://github.com/karthik/wesanderson) package.^[Feel
free to replace the palette with your favorite Wes Anderson movie palette.
How about Moonrise Kingdom?]

```{r}
# install.packages("wesanderson")
library(wesanderson)

darjeeling_pal <- function(...) {
  scale_fill_manual(
    ...,
    values = wes_palette("Darjeeling2"),     
    na.value = "grey"
  )
}

# this option replaces the default palette for all future ggplots
options(ggplot2.discrete.fill = darjeeling_pal)
```


```{r, fig.cap = "Voting intentions and age.", fig.width = 6, fig.height = 3.2}
ggplot(Chile2, aes(age, fill = vote)) +
  geom_density(position = "fill", col = "white")
```

This is probably quite fine as a comparison between the different
voting categories, but does make it very hard to see the *actual* age
distribution within each category. An alternative could be to use
the [ggridges](https://CRAN.R-project.org/package=ggridges) package, which
lets us plot shifted density plots.

```{r, fig.cap = "Distributions of age and voting intentions.", fig.width = 6, fig.height = 3.4}
# install.packages("ggridges")
library(ggridges)

ggplot(Chile2, aes(age, vote, fill = vote)) +
  geom_density_ridges2() +
  theme(legend.position = "none") # we don't need the legend here
```

All of the distributions are to some extent right-skewed^[The nomenclature for
skewness is confusing. When a distribution is said to be skewed to the
right, it means that the *tail* is on the right side, even though
it would probably make more sense to refer to the center of the distribution
(and say that it was then left-skewed), but we need to stick to convention
here. For more information, go to <https://en.wikipedia.org/wiki/Skewness>.]
The problem with this plot, however, is that it does not really answer the
question that we're interested in, which is to try to sift out to
which degree various voter groups voted for the different options.

Taken together, it's clear that those who are older are more prone to vote
yes for another term form Pinochet, while no voters as well as people
who intend to abstain are more frequent among the younger voters. One thing
that is devious about these visualizations is that the missing data seems
more prevalent from the plot types than is in fact the case. In the first plot,
this happens because the `NA` values are relatively more frequent among
the older ages and this part of the chart is enlarged in the proportional
version of the chart due to fact that this age category is in minority. In
the second, **ggridges** version, this happens simply because each geom
shows the distribution within each category and all sum up to the same
area.

## Age and Sex

We now continue to build our plot by introducing information on sex
as well. The best way to do this is to use facets. When you are faceting
on a single variable, you should mostly use `facet_wrap()` since it
works better than `facet_grid()` when there are many levels.

```{r, fig.cap = "Voting intentions across the ages and sexes.", fig.width = 7}
ggplot(Chile2, aes(age, fill = vote)) +
  geom_density(position = "fill", col = "white") +
  geom_rug(alpha = 0.05) + # display marginal distribution
  facet_wrap("sex")
```

Looking at this plot, it seems like there are a few differences between men
and women, for instance that more women then men seem poised to vote
yes among the younger population.

Note that we've
also included a rug geom in our plot (via `geom_rug()`), which displays
the *marginal* distribution for sex. Rugs (or other geoms that display
marginal distributions) are a good idea when you use proportional density
plots since it indicates for the reader (and you)
when they (or you) should be skeptical about the accuracy of the results.

We can take this even one step further, by involving our variables
on education, income, or population. Let's see if education matters.

```{r, fig.cap = "Voting intentions across age, sex, and educational attainment.", fig.width = 8}
Chile2 %>%
  drop_na(education) %>% # very small groups
  ggplot(aes(age, fill = vote)) +
  geom_density(position = "fill", col = "white") +
  geom_rug(alpha = 0.05) +
  facet_grid(sex ~ education)
```

Education certainly seems to matter for voting intentions. You can
for instance spot this among the youngest voters, where those with a
secondary level of education seems less prone to vote yes and more uncertain
too. Interestingly, the addition of education also seems too exacerbate
the differences between the sexes in the younger population.

One thing we really need to be very careful about is that we now have very
little data in some regions of the plot, notably the older age groups
that have post-secondary education (especially the women),
which seem to be very few by considering
the rug plot. For an even more clear demonstration of this, see the
following plot where we've switched to a stacked non-proportional density
plot.

```{r, fig.cap = "Voting intentions across age, sex, and educational attainment. A standard stacked density plot.", fig.width = 8}
Chile2 %>%
  drop_na(education) %>% # very small groups
  ggplot(aes(age, fill = vote)) +
  geom_density(position = "stack", col = "white") +
  facet_grid(sex ~ education)
```

This means that we should think again about our design. One possibility may be
to cut age into age groups before plotting or to limit the range of data
to regions where there are many observations.

## Satisfaction with the Status Quo

The survey also included a question on satisfaction regarding the
current state of affairs, which is recorded in the `statusquo` variable.
Let's take a look at this variable across age and region with a simple scatter
plot.

```{r fig.cap = "A simple (but unsuccessful) attempt to plot the satisfaction with the status quo in Chile against age and region."}
Chile2 %>%
  ggplot(aes(age, statusquo)) +
  geom_point(alpha = 0.5) +
  facet_wrap("region") +
  ylab("Support for the Status Quo")
```

The results are not attractive. There clearly appears to be some kind of
relationship between age, location, and support for the status quo, but
this plot fails to make those relationships clear, despite trying to mitigate
some of the issues of overlapping with opacity.

Our next attempt is to try a hexagonal binning plot instead.

```{r, fig.cap = "A hexagonal binning plot for the Chile survey data.", fig.width = 7, fig.height = 4.8}
Chile2 %>%
  ggplot(aes(age, statusquo)) +
  geom_hex(bins = 10) +
  facet_wrap("region") +
  ylab("Support for the Status Quo")
```

This looks a little better. We now see that there appears to
be a difference between the various regions and age groups. Support for the
current regime seems to be weak in the City of Santiago, particularly for
the younger age group. But there are at least two issues with
these hexagon plots.

One issue is that they put equal emphasis
on both of the variables mapped to the x and y axes. When we are primarily
interested in one of the variables---the support for the status quo here---this
does not communicate this message effectively. To highlight the distribution
of the status quo support, instead, it might therefore be more appropriate
to use a different type of design.

But most importantly, the measure of interest here is really the number of
observations in each bin as a proportion of the number of observations
in each region (facet). As it stands, the regions where there are few
observations now have basically all their bins colored in a single color,
which is not optimal. We could fix this relatively easily, but because
of the previously mentioned issue, we'll try something different.

What we'll do is to go back to the ridged density plots. We cut the
age variable into categories before plotting. To top it off, we
also try a more minimalist theme (`theme_minimal()`). Finally,
please make note of the `labeller` argument in `facet_grid()`, which is
important to remember whenever you have factor levels that are not
self-explanatory (as here).

```{r, fig.width = 5, fig.height = 7, fig.cap = "Gaussian density plots of satisfaction with the status quo in Chile in 1988 separated on age and region."}
Chile2 %>%
  drop_na(age) %>%
  mutate(`Age group` = cut_interval(age, 4)) %>%
  ggplot(aes(statusquo, region, fill = region)) +
  facet_wrap(~`Age group`, nrow = 4, labeller = label_both) +
  geom_density_ridges() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    y = "Region",
    x = "Satisfaction with the Status Quo"
  )
```

The plot is now a piece of work that we should be happy with.
Yes, we did lose some information when we
cut the age variable into categories, but the final plot does a much better
job of highlighting the goal of the visualization, which is to show
satisfaction with the current government in Chile ahead of the upcoming
plebiscite.

There remains some possible issues here. One is the use of density estimates
here, which may not be appropriate for all of these categories, particularly
not the *Metropolitan Santiago* region since it is small. An alternative could
be to switch to histograms instead.^[See
<https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html>
for a guide on how to use histograms instead of density plots.]

## Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-chile-plebiscite.Rmd>.
