---
title: "Worked Example: Air Quality"
author: "Johan Larsson"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# Air Quality in New York

As soon as you're faced with two numeric variables, your first instinct should
always be to want to put them on a scatter plot. One such situation could arise
with a data set like the `airquality` data set, which consists of
air quality measurements from New York, in May to September 1973.

The variables in this data set are

`Ozone`
: Mean ozone in parts per billion from 1300 to 1500 hours at Roosevelt Island

`Solar.R`
: Solar radiation in Langleys in the frequency band 4000–7700 Angstroms from
  0800 to 1200 hours at Central Park

`Wind`
: Average wind speed in miles per hour at 0700 and 1000 hours at LaGuardia
  Airport

`Temp`
: Maximum daily temperature in degrees Fahrenheit at La Guardia Airport.

`Day`, `Month`
: Self-explanatory.

Here are the first few rows of the data set.

```{r}
head(airquality)
```

Let's begin by just studying how wind affects air pollution (the level of ozone)
in the air. Take a minute to consider how you would map the variables to the
axes to study this relationship.

Well, our layman intuition tells us that it's likely that the wind affects air
pollution, and not the other way around, so the natural mapping here is
to place wind speed on the x axis and ozone on the y axis.

```{r, fig.cap = "Ozone and solar radiation from September to May in New York."}
library(tidyverse)

airquality %>%
  ggplot(aes(Wind, Ozone)) +
  geom_point() +
  labs(x = "Wind Speed (mph)", y = "Ozone (ppb)")
```

There is clearly a strong negative relationship here between wind speed and
ozone---with increasing levels of speed, air quality seems to improve. That's
hardly surprising if you think about it. But what kind of relationship is it?
Describing it as a linear relationship is not accurate.

The best way to describe this relationship is as an *inverse* relationship. That
is, ozone is inversely proportional to wind speed, which is equivalent to
saying that the relationship could be described as
$$
\text{ozone} = \frac{c}{\text{wind speed}}
$$
for some choice of a constant $c$.

Take some time to plot the associations between `Ozone`, `Solar.R`, `Wind`,
and `Temp` and try to think about what kind of patterns (if any) it is you
see.

## Longitudinal Data

This data is longitudinal in nature so it seems reasonable to want to
see how these measures vary across time. First, however, we need to
convert the day and month variables into a single variable representing time.
Here we use the handy `make_date()` function from the **lubridate** package.

```{r}
library(lubridate)

airquality_formatted <-
  airquality %>%
  as_tibble() %>% # tibbles are nicer to work with than data frames
  mutate(Date = make_date("1973", Month, Day))

airquality_formatted
```

Now we're ready to play with the data. The main interest of the data set is
of course air quality, so we want to look at ozone levels across time.

Here is a first attempt. We use the line geom, which is a natural choice
for this type of data.

```{r, fig.cap = "Air quality (ozone levels) in 1973", fig.width = 5.5, fig.height = 1.8}
p1 <-
  airquality_formatted %>%
  ggplot(aes(Date, Ozone)) +
  geom_line() +
  labs(y = "Ozone (ppb)")

p1
```

The result is possibly not quite as nice as you'd expect. Can you figure out
why there are multiple holes in the plot?

The reason is that there are missing values in the data set. This would not
have to be a problem for the visualization, weren't it for the fact that
observations that have neighboring observations that are missing also disappear
completely from the visualization when we use the line geom.

One solution in this case is to drop the missing values, which makes the
points connect.

```{r, fig.cap = "Air quality (ozone levels) in 1973", fig.width = 5.5, fig.height = 1.8}
airquality_nona <- drop_na(airquality_formatted, Ozone)

p2 <-
  airquality_nona %>%
  ggplot(aes(Date, Ozone)) +
  geom_line() + 
  labs(y = "Ozone (ppb)")

p2
```

One possible problem with this is that it's hard to tell which ranges of dates
correspond to missing values. One way to get around this is to add a point
geom too, which makes this clear.

```{r, fig.cap = "Air quality (ozone levels) in 1973.", fig.width = 5.5, fig.height = 1.8}
p3 <- p2 + geom_point()
p3
```

It seems quite clear that the air quality takes a turn for the worse during the
summer months.

### Air Pollution across the Week

One thing that is currently missing from the visualization but would shed
additional light on the relationships is the relationship between air pollution
and the type of day, that is, whether it is a weekday or a weekend. A possible
solution for this might be to color the weekend days differently, but here
we'll instead use the grid and axis ticks to display this information by
specifying scales with `scale_x_date()` and using `"weeks"` as input to the
`minor_breaks` argument.

```{r, fig.cap = "Air quality, in terms of ozone levels, in 1973", fig.width = 5.5, fig.height = 1.8}
p4 <- p3 + scale_x_date(minor_breaks = "weeks")
p4
```

It's unfortunately hard to see a clear pattern here, likely because
air pollution varies significantly from day to day due to other factors as well.
A possible way to solve this is to form a new variable, `Weekday`, in our
data set, which represents the weekday. We use the `wday()` function
from lubridate for this purpose.

```{r, fig.cap = "Air pollution across the week from May to June in 1973"}
airquality_1973 <-
  airquality_nona %>%
  mutate(Weekday = wday(Date))

airquality_1973 %>%
  ggplot(aes(Weekday, Ozone)) +
  geom_point() + 
  labs(y = "Ozone (ppb)")
```

Great, but now the point geom is probably not the best candidate for this job,
at least not without jittering and/or opacity. As an alternative, we could
instead use box plots here. To use box plots here, we need to also convert the
`Weekday` variable to a factor variable. To make sure that its order is also
kept, we use `as.ordered()` here. We do this directly in the call to
`ggplot()`, but note that this means that you need to fix the label of the
axis, which will otherwise turn out as `"as.ordered(Weekday)"`.

```{r, fig.cap = "Air pollution across weekdays from May to June in 1973", fig.width = 5.5}
airquality_1973 %>%
  ggplot(aes(as.ordered(Weekday), Ozone)) +
  geom_boxplot() +
  labs(x = "Weekday", y = "Ozone (ppb)")
```

Now we actually *can* see that there is a slight improvement in air quality over
the weekends, at least in terms of the median. At this point it would be 
interesting to carry on our investigation by looking at the outliers in this
plot to see if they correspond to particular events in the calendar.

Take some time to visualize the relationships between time and the other
variables in the data set too.

# Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-airquality.Rmd>.
