---
title: "Worked Example: Cars"
author: 
- "Johan Larsson"
- "Behnaz Pirzamanbein"
- "Jonas Wallin"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# Installing Tidyverse

If you haven't done so yet, you need to install the tidyverse collection
of packages. To do so, simply call the following line.
(Note that this will install several R-packages and
take a lot of time if you haven't already installed these packages.)

```{r, eval = FALSE}
install.packages("tidyverse")
```

If you already have **tidyverse** installed, you can skip this step. As in the
previous example, we will load the package with `library()` before using it.

# Importing Data into R

R is extremely versatile when it comes to handling data of different
formats and types. Much of this functionality is made available via R-packages,
which let you import data into R from Microsoft Excel, Stata, SAS, or SPSS files in
addition to standard file types such as comma-separated files (`.csv`)
and tab-separated files (`.tsv`). In this course, most of the data sets we use will be
available directly through R and R packages, but knowing how to import data directly is a
useful skill.^[And this might very well come in handy for your project, where you will
choose your data yourself.]

First we need to find some data to import. Download the US Cars dataset
that we have provided in the git repository for the course. The
data is available
[here](https://raw.githubusercontent.com/stat-lu/dataviz/main/data/us_cars.csv)
(this link may not work if you're browsing this page from
inside Canvas). Alternatively, you can call following lines of code to
download the dataset.

```{r}
# check that we have a data folder in our working directory, and if not create
# one
if (!dir.exists("data")) {
  dir.create("data")
}

# download the dataset to data/us_cars.csv
download.file(
  "https://raw.githubusercontent.com/stat-lu/dataviz/main/data/us_cars.csv",
  file.path("data", "us_cars.csv"),
  mode = "wb"
)
```

Now we load this dataset into R via the **readr** package (part of tidyverse).
This is a comma-separated file (note the file extension), so we use the
`read_csv()` function^[There is a `read_csv2()` function as well, which is for
semicolon-separated data.]

```{r}
library(tidyverse)
us_cars <- read_csv(file.path("data", "us_cars.csv"))
```

When you call `read_csv()`, the **readr** package helpfully prints a message to your console
with information about how the columns in the dataset were formatted when you
imported the data. Take a look at this information---does everything look
okay?

An alternative to this approach is to use **readr** to read data directly
from an URL into R. To do so, you simply use the URL instead of the file name
in the call to `read_csv()`, like this.

```{r}
us_cars <- read_csv(
  "https://raw.githubusercontent.com/stat-lu/dataviz/main/data/us_cars.csv"
)
```

## Taking a Glimpse

Whenever you start working with new data, always start by taking a look at the
data in raw form. The best first step is usually just to print the data to the
console or by calling `head()` (to see the first few lines).

```{r}
us_cars
```

If you want to see more rows of the data than are printed by default, try to call
`print()` with the `n` argument set to the number of rows you want, or use
`head()` in the same way.

Another useful function, particularly when there are many columns in the
dataset, is `glimpse()`.

```{r}
glimpse(us_cars)
```

# Wrangling

## Manipulation

In the tidyverse vocabulary, *filtering* refers to the process of selecting a
subset of rows (observations if the data is tidy) from your dataset, whereas
*selecting* means selecting a subset of columns (variables if the data is tidy).
We use the aptly named `filter()` and `select()` respectively for these tasks.

These functions return a transformed dataset. They do not change the object
that we pass to them. To keep a result for later use, we need to assign it to
an object using `<-`, just as we did when importing the data. We will see an
example of this below when we create a new variable with `mutate()`.

## Filtering

`filter()` is used on a dataset (tidyverse functions typically always
take data as its first argument) together with a number of logical expressions
that specify which rows to keep in the dataset. Recall that a logical expression
is a binary expression that relates the left-hand side to the right-hand side
in some way, for instance to check equality or inequality, like so:

```{r}
c(1, 2, 3) < c(0.2, 0.5, 3.8)
1:3 == 3:1
c("a", "b", "c") %in% c("a", "c")
```

Let's say that we only wanted to look at cars from Tennessee from years
2015 and onward. Then we can use `filter()` in the following way.

```{r}
filter(us_cars, state == "tennessee", year >= 2015)
```

## Selecting

If you think of filtering as slicing a dataset horizontally, `select()` does the
opposite, slicing a dataset *vertically*, by selecting a subset of the columns.
The interface is similar to `filter()`'s. You begin with the data and then
through various arguments select which columns it is you want to keep.

There is a plethora of possibilities when it comes to `select()`. We'll try to
cover some of the most common ones here, but see the documentation for
`select()` if you want to learn more.

The simplest option is simply to list the columns you want to keep.

```{r}
select(us_cars, brand, "model") # notice that you may omit quotation marks
```

When using `select()`, it's actually possible to change the names of the columns
by using a `new_name = old_name` pair, like this:

```{r}
select(us_cars, vehicle_id = vin, "price_in_dollars" = price)
```

If you instead want to drop a particular column, you just preface it with `-`
or `!`.

```{r}
select(us_cars, !brand, -model)
```

Notice that, when you only have negative indexing, the function assumes that you
want to keep all of the remaining columns.

It can, however, be tedious to manually select every column, in which case you
may use the `:` operator to specify a range instead.

```{r}
select(us_cars, title_status:state)
```

Finally, it can sometimes be useful to match columns by name in some way, for
instance if a number of columns that you want to contains a specific word. To be
able to do this, **tidyr** provides a set of helper functions, such as

* `starts_with()`,
* `ends_with()`, and
* `contains()`.

The manual entry for `select()` contains several examples using these helper
functions, as does the individual entries for each function.

Consider taking a little time to play around with `select()` and its helper
functions on the `us_cars` dataset.

## Mutating

Frequently when visualizing data you will want to transform that data in some
way. Perhaps you're more interested in the proportion than a number, want to
convert a value to a different unit or you just need to change the names of some
factor variables. In this cases, `mutate()` is your best friend.

Let's start by example. Say that we want to convert the price of the cars
in the `us_cars` dataset to Swedish kronor instead. Here's how to do this:

```{r}
mutate(us_cars, price = price * 8.92)
```

In the returned dataset, `price` has been replaced by the converted values.
The original `us_cars` object is unchanged, since we have not assigned the
result to an object. We can also create a new variable and save the result,
keeping the original prices alongside the converted prices.

```{r}
us_cars_sek <- mutate(us_cars, price_sek = price * 8.92)
select(us_cars_sek, price, price_sek)
```

We can now use `us_cars_sek` whenever we need the converted prices. The original
`us_cars` still contains the data that we imported, and we will continue to use
that object in the examples below.

Inside `mutate()`, you can basically use any operation that would work on a
typical vector in R. You can, for instance, convert between different vector
types, or use arbitrary functions (as long as they return a new vector).

```{r}
mutate(
  us_cars,
  year = as.integer(year), # convert year from a double to an integer
  state = toupper(state), # capitalize state names
  price = round(price, -2), # round to hundreds
  brand = fct_lump_prop(brand, 0.1) # lump together low-frequent brands
)
```

## Rename

When you produce visualizations, you need to make sure that annotation in the
plot, such as the axis titles, are legible. Many datasets, however, have
variable (column) names that are not. You *can* always fix this when you
set up your plot, but if you're producing visualizations it may become
tedious to rename the axes with every plot.

That is why it is often useful to rename your variables during the
data wrangling step. This also has the benefit of making your data wrangling
steps more readable. There are multiple ways to rename variables with
the tidyverse approach. We've already seen that you can use `select()` to
rename variables, but if you only want to rename and not select, you should
instead use `rename()`. Just as with `select()`, you use a
`new_name = old_name` pair to do so. Let's rename "vin", which stands for
vehicle identification number, to "vehicle_id".

```{r}
rename(us_cars, vehicle_id = vin)
```

You can have spaces in your variable names if you want to, but this is bad
practice because including special characters
like `%`, `&`, `$` in variable names can have undesired side-effects that you
do best to avoid.

# Grouping and Summarizing

Another common operation in data-visualization is to group and summarize data.
This is useful when you want to visualize a summary statistic rather than the
raw data. Before taking this route in data visualization, however, make it a
habit to ask yourself what you want the plot to show. Individual observations
help us see variation and unusual values, while summaries help us compare
features such as group means or medians. Sometimes it is useful to show both.
When we use summaries, we should also keep track of how many observations
contribute to each one.

## Grouping

To group data using the tidyverse methodology, we use `group_by()`
(from **dplyr**). This is a very simple function. You simply list all the
variables (columns) that you want to group by. Let's group the `us_cars` dataset
by country (USA or Canada).

```{r}
group_by(us_cars, country)
```

Okay, so not much happened! If you look closely, however, you'll see that this
tibble is now **grouped**---but what does this mean? Well, the truth is that
grouping is only useful when combined with functions that modify the data,
especially `summarize()`.

## Summarizing

`summarize()` takes a dataset---usually a *grouped* dataset---and a set of
functions that takes vectors as input and returns summary statistics,
such as

* `mean()`,
* `median()`,
* `sum()`, or
* `quantile(..., probs = 0.25)`.

Let's see how this looks in practice by computing the median and mean prices
of cars in USA and Canada.

```{r}
us_cars_grouped <- group_by(us_cars, country)

summarize(
  us_cars_grouped,
  median_price = median(price),
  mean_price = mean(price),
  number_of_cars = n() # captures the size of the group
)
```

Perfect! Note the use of the `n()` function that counts the number of
observations in each group. In this case, it highlights the problem that
can occur when we group, summarize, and plot, namely, that we lose information
about the size of the groups. It would not be reasonable to
draw conclusions about cars from Canada based on only seven observations.

# Missing Data

Missing data is an issue that is prevalent in much of the real-life data
that you may encounter, particularly when it comes to data involving
people. Dealing with missing data is an important topic but, as we said in
the introduction, it is a topic that lies mostly
outside the scope of this course. We do, however, recommend that you always take
time to consider the reasons for why data may be missing and for the
most part make it a habit to include missing data in your visualizations
instead of simply removing it.

What you do need to know, however, is that missing data (coded as `NA` in R)
may interfere with your work in R. Let's say that you're working with
sleep data on mammals using the `msleep` dataset from **ggplot2**, and want
to compute the mean number of hours in REM sleep:

```{r}
mean(msleep$sleep_rem)
```

The result here, `NA`, is a consequence of the presence of `NA` values in
the column `sleep_rem` in `msleep` and this behavior is actually the default
for many functions in R. We can use `is.na()` to check which values are
missing. It returns `TRUE` for a missing value and `FALSE` otherwise, so
summing the result tells us how many values are missing.

```{r}
sum(is.na(msleep$sleep_rem))
```

To compute the mean of the observed values, we can set `na.rm = TRUE` in the
call to `mean()`.

```{r}
mean(msleep$sleep_rem, na.rm = TRUE)
```

This excludes the missing values from this particular calculation. It does
not remove any rows from `msleep`, so the full dataset is still available
for other calculations and visualizations.

Alternatively, we can use `drop_na()` to create a dataset containing only the
rows where `sleep_rem` is observed.

```{r}
msleep_rem <- drop_na(msleep, sleep_rem)
nrow(msleep_rem)
mean(msleep_rem$sleep_rem)
```

The two means are the same because they use the same observed values. Notice
that we specified `sleep_rem` in the call to `drop_na()`. If we leave out the
column name, the function checks all columns and removes any row that has a
missing value in at least one of them.

```{r}
nrow(drop_na(msleep))
```

Why are fewer rows left this time? Some mammals have an observed value of
`sleep_rem` but are missing information in other columns. We don't need to
remove those rows to calculate the mean REM sleep time. Neither approach
tells us why values are missing or whether the observed mammals are
representative of those with missing values.

# Pipes

The pipe operator, `%>%`, is a staple in data science workflows because
it makes the process of managing data much more manageable, modular, and
readable. The pipe operator can be hard to get to grips with at first, but
once you do, you can write very clean code.

The pipe operator takes an object on its left-hand side (or line above)
and *pipes* it through to an object (almost exclusively a function) on the
right-hand side. The object on the left-hand side will enter
the function on the right hand side as *the first argument of that function*,
replacing (pushing back) any other argument that's already there.

Let's say that we have the following function, `my_frac()`, which takes
arguments `x` and `y` and returns `x/y`.

```{r}
my_frac <- function(x, y) {
  x / y
}
```

Now we can use this function on two numbers, like so:

```{r}
my_frac(2, 5)
```

If we were to do this with the pipe operator instead, it would look like this:

```{r}
# use the pipe operator
2 %>% my_frac(5) # equivalent to my_frac(2, 5)

# or we can do
2 %>%
  my_frac(5)
```

You're probably thinking that this doesn't seem very useful at all, given
that we've now spent more code to accomplish exactly what `my_frac()`
already accomplished very well without the pipe. Where the pipe operator
comes into its own, however, is when we need to chain multiple functions.

## Case Study using Pipes

Let's say that we want to take our `us_cars` dataset and perform a set of
operations, namely:

1. keep only black cars,
2. group cars by `state`,
3. compute the mean mileage and the number of cars per state,
4. sort the states by mean mileage, and
5. plot a simple bar chart of state versus mean mileage.

We will first compare three ways to prepare the summary table, and then use
it to make the plot:

1. use intermediary objects,
2. use function composition, or
3. use pipes.

## Intermediate Objects

To solve this by storing temporary objects, here's what we would do:

```{r}
black_cars <- filter(us_cars, color == "black")
black_cars_by_state <- group_by(black_cars, state)
black_cars_state_mileage <-
  summarize(
    black_cars_by_state,
    mean_mileage = mean(mileage),
    number_of_cars = n()
  )
black_cars_state_mileage <- arrange(black_cars_state_mileage, mean_mileage)
black_cars_state_mileage
```

The largest downside to this approach is that you're cluttering
the workspace with names of intermediary objects that you need to keep track of
and name with meaningful names, or alternatively name by incrementing a
counter such as `cars1`, `cars2`, etc. This latter approach is actually
what you'll often end up doing, and (at least for me) often leads to
mixing up the counters and ending up with the wrong results.

Storing intermediate results is sometimes precisely the right way to go
about this, especially when you will later use the intermediary result,
but in this case it leads to code that is hard to read and manage.

## Function Composition

An alternative is to use composite functions for our call. The code
above then looks like this.

```{r}
arrange(
  summarize(
    group_by(
      filter(us_cars, color == "black"),
      state
    ),
    mean_mileage = mean(mileage),
    number_of_cars = n()
  ),
  mean_mileage
)
```

Code like this is incredibly hard to manage (and read). Simply keeping
track of which argument belongs to which function is strenuous in itself.
On top of that, imagine that you want to swap order of one of these operations
or add another step somewhere, and it's easy to see why this is not an
attractive approach.

## Pipes

With pipes, we get the following result:

```{r}
black_cars_state_mileage <- us_cars %>%
  filter(color == "black") %>%
  group_by(state) %>%
  summarize(mean_mileage = mean(mileage), number_of_cars = n()) %>%
  arrange(mean_mileage)

black_cars_state_mileage
```

The code is clean and expressive. If we want to switch the order of the steps
or add another, it's only a matter of adding or removing a row of the
code. On top of that, you don't need to name any intermediate objects (but
remember that doing so is sometimes actually desirable).

The **tidyverse** is designed specifically with the pipe operator in mind. The
reason that it works so well with the tidyverse is that virtually all of the
main functions in the tidyverse take a data set as the first argument, which
means that piping always works. The base R functions and many other packages
out there don't necessarily abide by this principle, however, so make sure you
know what your functions are doing when you're using pipes or you may end up
with unexpected (and undesirable) results.

## Plotting the Result

Now we can use the summary table to make the bar chart. We use `reorder()` to
put the state labels in order of mean mileage. Sorting the rows with
`arrange()` alone does not set the order of the labels in the plot.

```{r cars-mileage-plot, fig.width = 7, fig.height = 8, fig.cap = "Mean mileage of black cars by state. Labels show the number of cars in each group."}
ggplot(
  black_cars_state_mileage,
  aes(x = mean_mileage, y = reorder(state, mean_mileage))
) +
  geom_col() +
  geom_text(aes(label = paste0("n = ", number_of_cars)), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "Mean mileage", y = "State")
```

The labels show how many cars contribute to each mean. Which states have only
one car? How much confidence would you have in comparing those states with
states that have many observations? Remember that these summaries describe
the black cars in this dataset; they do not automatically describe all cars
in each state.

Note that the pipe operator we have covered so far (`%>%`) is not part of
the base R distribution. Calling `library(tidyverse)` or `library(dplyr)` will
get you access to it.

## The Native Pipe Operator `|>`

R also has its own native pipe operator, available since version 4.1, which
you don't need any additional packages to use. It is written as `|>`. The
first two lines of our pipeline can also be written as follows.

```{r}
us_cars |>
  filter(color == "black")
```

For all purposes that we use it for in this course, you may as
well use the native pipe operator if you prefer to. `%>%` has a bit of
additional functionality, but we will not need it in this course.

If you want to read more about pipes or just need a longer introduction, we
recommend to take a look at [the piping
section](https://r4ds.had.co.nz/pipes.html) of the R for Data Science book.

Have fun piping in the tidyverse!

# Pivoting

One of the most important data wrangling tools when it comes to data
visualization is *pivoting*, which can be used to transform messy data into
tidy data, which in turn will make visualization much easier for us.

Let's see what this can entail by looking at a simple dataset, `table2` from
the **tidyr** package, which contains information on tuberculosis cases
in a few countries from 1999 and 2000.

```{r}
table2
```

This data is *not* tidy---but why not? Take a moment to consider this for
yourself before moving on.

Did you figure it out? The problem is that `cases` and `population` are two
different variables but don't have their own separate columns. To fix this, we
need to reshape the data by pivoting it to a wider form using `pivot_wider()`.
Before you continue, take a look at the documentation for the function by
calling `help(pivot_wider)` (or `?pivot_wider`) in R and see if you can make
sense of the manual entry for the function.

The function has many arguments but we only need to concern ourselves
with `data`, `names_from`, and `values_from` right now. `names_from` should
indicate which columns store the names of the variables we want to pivot,
while  `values_from` should contain those variables' values. Putting this
together, we get the following.

```{r}
table2_tidy <-
  pivot_wider(
    table2,
    names_from = "type",
    values_from = "count"
  )

table2_tidy
```

Now it's easy to visualize this data!

```{r, fig.cap = "Tuberculosis cases in Afghanistan, Brazil, and China in 1999 and 2000"}
ggplot(table2_tidy, aes(year, cases, fill = country)) +
  geom_col(position = "dodge")
```

Sometimes it's necessary to pivot in the other direction: from wide to long, in
which case we need to use `pivot_longer()` instead. In this example, we can use
it to recover the original country-year-type observations. Pivoting back and
forth does not always recover the original data, for instance if values were
aggregated when pivoting to a wider form. Just like you did for `pivot_wider()`,
read the documentation for `pivot_longer()` and then see if you can figure out
by yourself how to pivot `table2_tidy` back into the same long format as `table2`.

When we use `pivot_longer()` we need to tell R what columns we want to stack,
which we do through the `cols` argument. Here we simply specify the
names of the columns (cases and population), and then use `names_to` and
`values_to` with whatever names we think are appropriate.

```{r}
table2_messy <-
  pivot_longer(
    table2_tidy,
    cols = c("cases", "population"),
    names_to = "type",
    values_to = "count"
  )
```

# Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-examples/cars.Rmd>.
