---
title: "Worked Example: First Steps with ggplot2"
author: "Johan Larsson"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# ggplot2

ggplot2 is part of the **tidyverse** so you can simply call

```{r}
library(tidyverse)
```

to load ggplot2.

# `ggplot()`

The main function of ggplot is the `ggplot()` function. It has two main
arguments that you need to be aware of:

`data`
: the (default) dataset to use for the plot

`mapping`
: the aesthetic mappings

When calling `ggplot()` we typically don't call these arguments by their name
and instead simply supply them in order.

## `data`

The `data` argument obviously needs a dataset and this should usually come in
the form of a `data.frame` or a `tibble`. Let's starts with a very simple data
set, `mtcars`, that contains a few variables related to fuel consumption
of cars.

Before doing anything else, always take a little time to look at bit of the
data. If the dataset has relatively few variables (less than ten), the `head()`
function is usually your best option; it simply displays the first few rows of
the data.

```{r}
head(cars)
```

The second argument of `heads()` decides how many rows to show. Modify this
value to show the first 10 cars of the dataset.

`glimpse()` is very useful when you have more variables (since it is more
compact). Call `glimpse()` on `cars` and notice that it lists the variables in a
table along with the first values of each variable.

## `mapping`

The `mapping` argument should be supplied with a call to the `aes()` function
that lists the mappings between variables in your data and aesthetics of the
plot. Which aesthetics you choose to map to depends on what geoms you intend to
use. But for now, let's assume that we want to map engine displacement (`disp`) to the x
axis, and fuel consumption (`mpg`) to the y axis. To do so, we simply use
`aes(x = disp, y = mpg)` as the input to the `mapping` argument. Because the
first two arguments of `aes()` are `x` and `y`, we can do without calling them
explicitly.

Let's put this together:

```{r blank-canvas}
ggplot(mtcars, aes(disp, mpg))
```

Why does this result in a blank canvas? After all, we did tell ggplot to map
engine displacement to the x axis and fuel consumption to the y axis.

The answer is that we still haven't told ggplot2 which layers we want to use.

# Layers

The aesthetic mapping is not enough by itself---we also need a layer to display
our data. In **ggplot2**, we accomplish this by adding (stacking) layers to the
output of `ggplot()`. There are several ways to construct layers in ggplot, but
the simplest and most common approach is to use the functions named `geom_*()`
(such as `geom_point()`). You can think of these functions as a short cut
to generating layers. Specify the geom you want and ggplot2 uses its 
predefined defaults to provide the other components of the layer.

To provide a layer with the point geom to the plot, you simply need to call
the following lines.

```{r cars-plot}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point()
```

So, in fact, `geom_point()` doesn't just return a point geom. Instead it returns
a complete layer, including a point geom.

Notice the use of `+` here to add layers to the canvas that we started out with.
Also observe that we didn't need to specify what data to use or which mappings
we wanted in the call to `geom_point()`---they were inherited from the
`ggplot()` call. If you want to avoid this type of inheritance (which you
will want when constructing plots from different data sources), you can
do so by specifying `inherit.aes = FALSE` in the call to `geom_point()`.

Alternatively, you can define the mappings in the `geom_point()` function
directly, which also has a `mapping` argument. Practically speaking, the
lines above turn out to be equivalent to those below.

```{r cars-plot-aes}
ggplot(mtcars) +
  geom_point(aes(disp, mpg))
```

This will come in handy for us later on in the course.

The layer functions, such as `geom_point()`, also have its own `data` argument,
which means that we can define the data directly in the `geom_point()` function
too:

```{r cars-plot-data}
ggplot() +
  geom_point(aes(disp, mpg), data = mtcars)
```

Again, we retrieve the same plot. But note that the `mtcars` data set
is now available only to the `geom_point()` layer. If we were to add
an additional layer, then that layer wouldn't know about the `mtcars` data
set (nor the aesthetics mappings defined in the `geom_point()`) call.

Now try to switch geom for your plot, for instance
by replacing `geom_point()` with `geom_line()`.

Now let's try using `geom_rect()` instead.

```{r cars-plot-fail, eval = FALSE}
ggplot(mtcars, aes(disp, mpg)) +
  geom_rect()
```

Well that didn't work, but why not? Well, the reason is simply that
`geom_rect()` uses the rectangle geom and a rectangle needs more aesthetics than
just x and y coordinates.

# Aesthetics

Look at the documentation of the `geom_point()` function. Under **Aesthetics**
we can see that this layer understands the following aesthetics:

- **x**
- **y**
- alpha
- colour
- fill
- group
- shape
- size
- stroke

The bold face of the first two aesthetics signify that they must be provided.
The remaining aesthetics have been supplied defaults, such as `alpha = 1` (no
transparency) and `colour = "black"` (black points). But these aesthetics can
also be changed, either by mapping them to another variable in our data or by
manually adjusting them directly.

When you change an aesthetic by specifying a value (and not a mapping) you
simply specify that aesthetic in the call to the geom you would like to change.
Let's change the color of the points to navy blue, increase the size, and change
shape. Doing so is a breeze with ggplot2:

```{r cars-navy}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point(
    colour = "navy",
    size   = 1.8,
    shape  = 4
  )
```

Take some time to play around with the various aesthetics. Switch geom and try
to flip the axes. Can you figure out what the `stroke` aesthetic does?

It might be worth noting that some of the aesthetics used here have synonyms
that you will likely come across every now and then, for instance `cex = size`
(for points), `col = color = colour`, and `lty = shape` (for points), which are
there mostly for reasons that are historical or have to do with compatibility.

# Other Ways to Generate Layers

So far we've learned that the most common way to generate layers is
to use `geom_*()` functions, but there are alternatives. In particular,
you can also generate layers from `stat_*()` functions. Recall that
the `geom_*()` functions generate layers through geom specifications. `stat_*()`
functions generate layers through statistical transformations, providing
defaults for the other features of a layers, such as the geom.

The `stat_identity()` function, for instance, specifies the "identity"
transformation, which is really just a function that does no transformation:
it simply returns its input as output. A natural default geom for the
identity transformation is a point geom, so if we use
`stat_identity()` like so:

```{r cars-stat}
ggplot(mtcars, aes(disp, mpg)) +
  stat_identity()
```

We get the same scatter plot as before.

The example above is not really of much practical use. But `stat_*()` functions
occasionally do come in handy. They are, for instance, natural to consider
when your goal from the outset is to visualize the result of
a statistical transformation (other than the identity function), such
as medians of some
distribution, for instance. In this example, we've used 
the `stat_summary()` function to compute medians of engine
displacement across all unique values of numbers of cylinders.

```{r}
ggplot(mtcars, aes(cyl, disp)) +
  stat_summary(fun = median, geom = "line") +
  stat_summary(fun = median, geom = "point")
```

# Limits

It is sometimes handy to be able to change the limits of the visualization,
in particular the limits for the x and y axes. You can do so either using
the `lims()` function or via the wrappers `xlim()` and `ylim()`, which modify
the limits for the x and y axis respectively.

This will come in handy when you for instance want to include a particular value
on the axes (such as zero) or just want to make room for annotation.
Here we've taken the previous plot and changed the limits to include
0 on the y axis. 

```{r}
ggplot(mtcars, aes(cyl, disp)) +
  stat_summary(fun = median, geom = "line") +
  stat_summary(fun = median, geom = "point") +
  ylim(0, 300)
```

# Scales (and Guides)

As you may recall, scales are the functions that take your data and
decide how it is to be displayed on the canvas.
The `lims()`, `xlim()`, and `ylim()` functions actually modify the scales
of the visualization. If you want more control over the scales, however,
you can achieve that using `scales_*()` functions.

Every ggplot must have a scale function attached to every aesthetic mapping.
The reason we haven't had to specify them before is, as you might have
guessed, that ggplot2 has filled in the blanks for us using reasonable defaults.

Look at the following plot.

```{r}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point()
```

It is actually equivalent to this one:

```{r}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  scale_x_continuous() +
  scale_y_continuous()
```

This makes sense. Speed and distance are continuous variables, so they
should have continuous scales.

`scale_*()` functions can be used to modify the scales in many ways. We can,
for instance, reverse the direction of the scale or use a logarithmic
version of it instead.

The scale functions in fact not only modify the scales themselves, but
also modify the *guides*. If you recall, a guide is the *inverse* of a scale:
it tells the reader how to make sense of scale. In the example above,
for instance, we have two guides: the two axes, which include all the tick
marks and numbers.

Here's an example where we've gone a bit crazy with these settings.

```{r}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  scale_x_reverse(
    breaks = c(5, 10, 15),
    minor_breaks = 7,
    position = "top"
  ) +
  scale_y_log10()
```

# Guides

For other guides, such as legends for sizes, shapes, and color mappings,
most of the control of the guides is instead covered through the `guides()`
function. In fact, it may be better to think of the `guides()` function
as a function that controls legends.

Most of the time, the only reason you'll have to invoke the `guides()` function
will actually be to *remove* guides that are redundant[^1]. To do so, just
set the respective guide to `"none"`, like this:

```{r}
ggplot(cars, aes(speed, dist)) +
  geom_point() +
  guides(x = "none")
```

If you want to read more about modifying guides (legends), check out
this page: <https://ggplot2.tidyverse.org/reference/guides.html>.

# Labeling Guides

How you label your guides matters a lot and it's probably the thing in the
course that we nitpick on the most in (save maybe for improper figure sizing
and non-descriptive figure captions). 

It's fortunately very easy to do in ggplot2. Just use the `labs()` function.
The basic recipe is `labs(<guide> = <name>)`. So, to label the guides
in the `mtcars` example, for instance, you would just do:

```{r}
ggplot(mtcars, aes(disp, mpg, color = factor(cyl))) +
  geom_point() + 
  labs(
     x = "Engine Displacement",
     y = "Fuel Consumption\n(Miles per Gallon)",
     color = "Number of\nCylinders"
  )
```

Notice the use of the special character `"\n"` in these labels. As you can
see on the plot, they add line breaks inside the titles, which is very useful
if you have long titles that don't fit otherwise or just don't want
to waste space on a long title label for the legend.

Just like with limits, there are shortcuts for the x and y coordinates here
too: `xlab()` and `ylab()`.


# Changing the Appearance of Your Plots

The appearance of ggplots can be varied in a great many ways and the interface
for doing so is the `theme()` function. The number of options is
staggering, but thankfully you will rarely need to mess with them directly,
except for a couple of options that will often come in handy.

In particular, it's very useful to know how to re-position legends,
which you do using the `legend.position` argument in the call. By default,
legends will always be placed on the right of the plot. But sometimes
you need that space for the canvas, perhaps just to alleviate issues with
overlap. It's very easy to do so:

```{r}
ggplot(mtcars, aes(disp, mpg, color = factor(cyl))) +
  geom_point() + 
  theme(legend.position = "bottom") # top, left, right, bottom
```

In our particular plot, we might even prefer to put the legend inside the plot
itself, given that there is ample space in the upper-right corner of the plot.
If you want to place the legend inside the canvas, then you supply a length-two
numeric vector. Each value needs to lie within [0, 1] and represent normalized
coordinates of the canvas. Placing legends inside the canvas well typically
comes down to trial-and-error and will depend on the size of the plot in the
output.

```{r}
ggplot(mtcars, aes(disp, mpg, color = factor(cyl))) +
  geom_point() + 
  theme(legend.position = c(0.8, 0.8))
```

The full list of options for the `theme()` function is available at
<https://ggplot2.tidyverse.org/reference/theme.html>. 

# Themes

If you want to modify the appearance of all of your plots, the simplest way of
doing so is to switch theme for your plots. There are many built-in themes in
ggplot2[^2]. The default theme, which you've seen exemplified several times by
now, is `theme_grey()` and for most of the plots in this course, we will stick
to that.

Changing the theme for your plots can be done on a plot-by-plot basis, by
simple adding the theme you want to your plot, like this:

```{r}
ggplot(mtcars, aes(disp, mpg, color = factor(cyl))) +
  geom_point() + 
  theme_bw()
```

But it's also possible to change the default theme for your entire session by
calling `theme_set()` somewhere in your R source (or R Markdown) document, like
this:

```{r}
theme_set(theme_minimal())

ggplot(mtcars, aes(disp, mpg, color = factor(cyl))) +
  geom_point() 
```

You are free to choose theme for yourself, but I highly recommend that you try
to stick to one and the same theme for any given report to create a uniform
design for your reports.

There are other packages that provide additional themes for ggplot2 too. If you
are interested, we recommend you check out
[ggthemes](https://cran.r-project.org/package=ggthemes) and
[cowplot](https://cran.r-project.org/package=cowplot).

# Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-examples/first-steps-with-ggplot2.Rmd>.

[^1]: You'll find out more about redundant ink later on in the course.
[^2]: For a full list, see
     <https://ggplot2.tidyverse.org/reference/ggtheme.html>.
