---
title: "Worked Example: Food for Thought"
author: "Johan Larsson"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE, message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# Budget Share of Food for Spanish Households

In this worked example we will try to make use of our newfound tools to handle
a big data set. The data set that we will be looking at consists of house
expenditures on food from Spanish households in 1980. The objective of our
study will be to explore how food expenditure is related to the various
other variables in the data set.

The data set is available in the
[Ecdat](https://CRAN.R-project.org/package=Ecdat) package. Start by installing
it (if you haven't already).

```{r, eval = FALSE}
# install the Ecdat package
install.packages("Ecdat")
```

Then load the data set directly into your workspace by calling the `data()`
function as we show next.

```{r}
data("BudgetFood", package = "Ecdat")
```

There are 23,972 observations and six variables in this dataset:

`wfood`
: Percentage of total expenditure which the household has spend on food

`totexp`
: Total expenditure of the household

`age`
: Age of reference person in the household

`size`
: Size of the household

`town`
: An ordered factor of the size of the town where the household is placed.
  1 for small towns and 5 for big ones.

`sex`
: sex of reference person (man, woman)

# Wrangling

Before we start visualizing this data set, we need to format it properly.
The names of the variables are fine, but the `town` variable
should be formatted as an ordered factor and `age` and `size` should 
be formatted as integers.

```{r}
library(tidyverse)

budget_food <- 
  BudgetFood %>%
  mutate(
    age = as.integer(age),
    size = as.integer(size),
    town = factor(town, ordered = TRUE),
  )
```

# Food and Total Expenditure

Let's start naively, trying out a simple scatter plot of food expenditure
versus total expenditure.

```{r foodexp-naive, fig.cap = "A scatter plot of total expenditure versus expenditure on food as a fraction of total expenditure"}
ggplot(budget_food, aes(totexp, wfood)) +
  geom_point() +
  labs(x = "Total Expenditure (Pesetas)", y = "Food Expenditure")
```

Despite the large amount of overlap, we can still glean a few insights. First
off, food expenditure as a proportion of total expenditure decreases with total
expenditure. This speaks to our intuition. Second, there is a number of
observations that report no food expenditure whatsoever. These observations are
likely outliers, perhaps people living in elderly care centers or something
similar. Let's remove these observations from the data set.

```{r remove-outliers}
budget_food_cleaned <- 
  budget_food %>%
  filter(wfood > 0)
```

At this point we could try to combat the overlap by reducing opacity, but
another trick that is sometimes overlooked is to instead try a scale
transformation. Instead of a linear scale for total expenditure, we could use a
logarithmic scale.

```{r logplot, fig.cap = "A scatter plot of total expenditure versus expenditure on food with total expenditure on a logarithmic scale"}
ggplot(budget_food_cleaned, aes(totexp, wfood)) +
  scale_x_log10() +
  geom_point() + 
  labs(x = "Total Expenditure (Pesetas)", y = "Food Expenditure")
```

But since we have just learned a few new tools, let's try to make use of them
now. In this case, we are going to try a hexbin plot.

```{r hexplot, fig.cap = "A hexbin plot of total expenditure versus expenditure on food as a fraction of total expenditure"}
ggplot(budget_food_cleaned, aes(totexp, wfood)) +
  geom_hex() + 
  labs(x = "Total Expenditure (Pesetas)", y = "Food Expenditure")
```

This plot definitely works better, showing the relationship in a clear fashion.

# Household Size

Now that we have managed to sort out of plot of food and total expenditure,
we'll see if we can find a relationship with the size of the household too.
Before we look at the complete picture, it is a good idea to first
check what the overall relationship between household size and food expenditure
is. To do so, we first construct a boxplot.

```{r boxplot, fig.cap = "A boxplot of food expenditure as a fraction of total expenditure versus household size"}
ggplot(budget_food_cleaned, aes(as.factor(size), wfood)) +
  geom_boxplot() + 
  labs(x = "Household Size", y = "Food Expenditure")
```

In general, food expenditure seems to be higher with larger households, which
is probably what one would expect. But this is not the case for households
with one or two members. Take a minute to figure out if that makes sense to you.

Let's return to our hexbin plot design, but now introduce household size
as facets.

```{r hexbin-size, fig.cap = "A hexbin plot of total expenditure versus expenditure on food across household sizes", fig.width = 8, fig.height = 6}
ggplot(budget_food_cleaned, aes(totexp, wfood)) +
  geom_hex() +
  facet_wrap("size", labeller = labeller(size = label_both)) + 
  labs(x = "Total Expenditure (Pesetas)", y = "Food Expenditure")
```

This plot looks promising but is crippled by the fact that some of the groups
are small, which creates problems for the hexbin plot. In fact, hexbin plots
are generally problematic when the number of observations if facets
differ because all the facets share the same scale (binning).

To get around this problem, we are going to lump all of the households of
size 6 or larger into one level, using `fct_other()`, a function from the
**forcats** package. We specify the levels we want to *keep* using the
`keep` argument, and all of the others are lumped together.

```{r hexbin-binned, fig.cap = "A hexbin plot of total expenditure versus expenditure on food across household sizes. The 6+ category represents household sizes of 6 or more people.", fig.width=8, fig.height = 5}
budget_food_binned <- 
  budget_food_cleaned %>%
  mutate(
    size = as.factor(size),
    size = fct_other(size, keep = c(1:5), other_level = "+6")
  )

ggplot(budget_food_binned, aes(totexp, wfood)) +
  geom_hex() +
  facet_wrap("size", labeller = labeller(size = label_both)) +
  labs(x = "Total Expenditure (Pesetas)", y = "Food Expenditure")
```

This is now starting to look quite interesting.

# Smoothers

As an alternative to the hexbin plots, we could instead use a smoother to look
at the relationship between total expenditure and food expenditure. 

Smoothers work particularly well when you expect there to be some kind of
causal relationship between two variables. They also help when you have a
smaller (but still sufficiently large) number of observations in the upper and
lower ranges of the variable on the x axis.

Let's try a smoother using a LOESS model, which makes few assumptions about the
data and is relatively robust to outliers.

```{r smoother, fig.cap = "A scatter plot of total expenditure, expenditure on food as a fraction of total expenditure, and household size. A LOESS smoother, including approximate 95% confidence intervals, is used.", fig.width = 8, fig.height = 4.5}
ggplot(budget_food_binned, aes(totexp, wfood)) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "loess") +
  facet_wrap("size", labeller = labeller(size = label_both)) +
  labs(x = "Total Expenditure (Pesetas)", y = "Food Expenditure")
```

The problem with this plot is of course that the smoothers are clearly
inappropriate for the extreme values of total expenditure. To work around
this problem, we can create a separate data set to use in the
`geom_smooth()` function only.

To do so, we create a new data set that we first group by the 
variable we are faceting on, `size`, and then filter the data set such
that only the observations of `totexp` (total expenditure) large than
the 0.05 quantile and smaller than the 0.95 quantile remain in the data set.

Then we use this data in the `data` argument to `geom_smooth()`.

```{r smoother-fixed, fig.cap = "A hexbin plot of total expenditure versus expenditure on food", fig.width = 8, fig.height = 4.5}
budget_food_smoother_data <-
  budget_food_binned %>%
  group_by(size) %>%
  filter(totexp > quantile(totexp, 0.05) & totexp < quantile(totexp, 0.95))

ggplot(budget_food_binned, aes(totexp, wfood)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "loess", data = budget_food_smoother_data) +
  facet_wrap("size", labeller = labeller(size = label_both)) +
  labs(x = "Total Expenditure (Pesetas)", y = "Food Expenditure")
```

Go ahead and see if you can figure out ways to improve this plot further.
Consider using the scale transformation we touch on earlier and see how
that works together with the smoothers and hex binning.

# Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-examples/food-for-thought.Rmd>.
