---
title: "Worked Example: Gapminder"
author: 
- "Johan Larsson"
- "Behnaz Pirzamanbein"
output:
  html_document:
    toc: true
    toc_float: true
    highlight: pygments
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# Gapminder

Do you remember the animated plots we produced in the introductory
lecture for this course based on the Gapminder Hans Rosling
animated visualization?

```{r rosling, echo = FALSE, out.extra='style="border: none;"', out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/jbkSRLYSojo")
```

In this worked example, we'll figure out how to
reproduce that plot as both an animated and an interactive visualization.

# Animated Visualization

The dataset that we'll use is available via the
[gapminder](https://CRAN.R-project.org/package=gapminder) package. So go ahead and install that package.

```{r eval = FALSE}
install.packages("gapminder")
```

This makes the dataset available in an object called `gapminder`. These are the first few rows of the dataset.

```{r}
library(gapminder)

head(gapminder)
```

The variables are self-explanatory.

Let's jump right in and create a bubble plot, faceted by year (which we have grouped), with population size mapped to the size of the bubbles, and GDP per capita and life expectancy mapped to the x and y axes, respectively.

```{r fig.cap = "Life Expectancy and GDP per capita from 1950 to 2010.", fig.width = 8, fig.height = 7}
library(tidyverse)

gapminder %>%
  mutate(years = cut_interval(year, length = 5)) %>%
  ggplot(aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap("years") +
  labs(
    y = "Life Expectancy",
    x = "GDP per Capita",
    size = "population"
  )
```

Just as we metioned in the first lecture, this visualization is not working out so well for us yet. Let's make it animated instead. For this, we'll use
the [gganimate](https://CRAN.R-project.org/package=gganimate) package.
First install the package.

```{r, eval = FALSE}
install.packages("gganimate")
```

To use the `gganimate` package you also need a *renderer* to produce animated images. You can use either [gifski](https://gif.ski/) or [ImageMagick](https://imagemagick.org/). We recommend the former, as `gganimate` defaults to gifski if it is installed, but either option will work just fine. Run one of the following lines, or both, to install a renderer.

```{r, eval = FALSE}
install.packages("gifski")
install.packages("magick")
```

We build the plot as before, but now we make it animated by adding
the `transition_time()` function to the plot and using the
title label to show the current year.

```{r, fig.cap = "GDP per capita and life expectancy for some of the countries of the world.", fig.width = 7, fig.height = 6}
library(gganimate)

ggplot(gapminder,  aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  labs(
    title = "Year: {frame_time}", # special glue syntax,
    y = "Life Expectancy",
    x = "GDP per Capita",
    size = "population"
  ) +
  transition_time(year)
```

If you think the plot is still crowded, we could alternatively use
facets to separate continents. Here we also make use of the
`country_colors` object that is included in the **gapminder** package.

```{r, fig.cap = "GDP per capita and life expectancy; now with facets!", fig.width = 7, fig.height = 6}
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(alpha = 0.5) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(
    title = "Year: {frame_time}",
    x = "GDP per capita",
    y = "Life expectancy"
  ) +
  transition_time(year)
```

So far, our plot effectively shows the trends across the world's continents, but it becomes difficult to use when focusing on a specific country. A remedy for this issue could be to use labels to identify which bubble corresponds to which country. However, the large number of countries means that labelling *all* of them may not be a particularly good idea.

Instead, we'll select the two largest countries (at the most recent timestamp) on each continent to label. First, we will store the names of these countries in a vector named, `large_country_names`.

The following steps first filter the dataset to retain only
observations from the most recent year, identified by `max(year)`. Next, the dataset is grouped by continent. It is then slices to keep only the countries with the largest and second-largest population (`pop`) values within each group (continent). Finally the country names are extracted using `pull()` function.

```{r}
large_country_names <-
  gapminder %>%
  filter(year == max(year)) %>%
  group_by(continent) %>%
  slice_max(pop, n = 2) %>%
  pull(country)

large_country_names
```

Then, we filter the original dataset to create a separate dataset specifically for our labels.

```{r}
large_countries <- filter(gapminder, country %in% large_country_names)
```

Now, we put everything together; we also adjust the *easing* of aesthetics from linear to cubic in-and-out using `ease_aes()`, to highlight that our data spans only 5-year intervals.
We use `geom_label_repel()` from the `ggrepel` package for labeling countries, which helps prevent label overlap.^[Working with labels and animated visualizations presents a challenge. Here, I had to fine-tune the settings, particularly `nudge_x` and `nudge_y`, several times in order to get something that looks good.]

```{r, fig.cap = "Life expectancy and GDP per capita with countries. The two largest countries by population at the start have been labeled.", fig.width = 8, fig.height = 7}
library(ggrepel)
library(scales)

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(alpha = 0.5) +
  geom_label_repel(
    aes(gdpPercap, lifeExp, label = country),
    inherit.aes = FALSE,
    seed = 1, # important when animating
    nudge_x = 6,
    nudge_y = -10,
    data = large_countries
  ) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12), labels = scientific) +
  scale_x_log10(labels = scientific) +
  facet_wrap(~continent) +
  labs(
    title = "Year: {frame_time}",
    x = "GDP per capita",
    y = "Life expectancy"
  ) +
  transition_time(year) +
  ease_aes("cubic-in-out")
```

Note that I used the `scales` package and set `labels = scientific` for the legend and the `x-axis` to shorten large numbers.

# Interactive Visualization

Interactive visualizations are particularly effective for exploring complex datasets like this one. We'll use the [plotly](https://CRAN.R-project.org/package=plotly) package, which integrates well with `ggplot`, as discussed in the lecture. Begin by installing the package.

```{r, eval = FALSE}
install.packages("plotly")
```

Then, we load the package.

```{r}
library(plotly)
```

Now, let's redraw the plot and add an interactive slider for year selection with `plotly`. Note the new `frame` mapping added to `geom_point()`. This special mapping informs `plotly` about the variable to use for separating the visualization into frames.

```{r, fig.cap = "An interactive visualization using plotly for the Gapminder data.", fig.width = 8, fig.height = 7}
p <- ggplot(gapminder, 
            aes(gdpPercap, lifeExp, size = pop, color = country))+
  geom_point(aes(frame = year), alpha = 0.5) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(x = "GDP per capita", y = "Life expectancy")

ggplotly(p)
```

Notice how seamless the transition from `ggplot` to interactive plots facilitated by `plotly`.

# Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-examples/gapminder.Rmd>.
9
