---
title: "Worked Example: Heavy Metal"
author: "Johan Larsson"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# Meuse River Data

In this worked example, we'll take a look at the *Meuse river data set*
from the [sp](https://CRAN.R-project.org/package=sp) package^[The sp package
is a full-featured solution for visualization data, and can be used in its
own right to create maps in R.]. 

You of course need to download and install the package if you don't have
it already. Then make the data set available using `data(meuse)`.

```{r}
library(sp)
data(meuse)
```

This map data is just a simple `data.frame`, but since we will be working
with simple features, we need to convert this data set to such. First, however,
notice that if
you look at the `x` and `y` values here, it's clear that they're actually
not longitude and latitude measurements.
By looking at the documentation for `meuse`, we instead see
that they are in fact Rijksdriehoek (RDH) (Netherlands topographical) map
coordinates. To everything work out correctly later on, we need to make
sure that our simple features data understands this information.

To convert the data into a simple features object, we use the
`st_as_sf()` function, using the `coords` argument to specify which
variables represent our x and y coordinates, and then we use the `crs`
argument to denote which projection system they are measured in.
RDH coordinates have the EPSG code 28992. So we use that as input.

```{r}
library(sf)

meuse_sf <- 
  meuse %>%
  st_as_sf(coords = c("x", "y"), crs = 28992) %>%
  st_transform(crs = 4326)
```

Before we dig into this data set, let's set up the ggplot2 theme for mapping.
This time, we use the `theme_map()` (instead of `theme_void()`, which we used
previously) from the [ggthemes](https://CRAN.R-project.org/package=ggthemes)
package.

```{r}
library(tidyverse)
library(ggthemes)

theme_set(theme_map(base_size = 11))
```

Let's start with a simple plot, looking at lead, creating a bubble plot by
mapping the presence of lead to size of the points.

```{r, fig.cap = "Concentration of lead around the river Meuse.", fig.width = 4, fig.height = 5}
ggplot(meuse_sf) +
  geom_sf(aes(size = lead)) +
  theme(legend.position = "right")
```

# Adding the River

The `meuse` data set is so called because it's related to pollution data
around the river Meuse, so it makes sense that we actually try to include
the location of this river in our plot too.

To do so, we need some map data that represents this river. Our trusty
**rnaturalearth** package yet again comes to the rescue, although this
time the data is actually not included directly in the package
but must instead be downloaded using the `ne_download()` function.

Let's begin by downloading the data.

```{r}
library(rnaturalearth)

rivers <- ne_download(
  scale = 10,
  type = "rivers_lake_centerlines",
  category = "physical",
  returnclass = "sf"
)
```

`rivers` contains rivers and lakes from all across the world, which means
that if we were to plot it, we would get a lot of information we don't need.
There are two ways around this, but either way we first need to figure out
a bounding box that includes the coordinates from the data set, perhaps
with a little extra margin. As I hope you've started to realize by now,
there often is a `st_*()` function for whatever mapping related issue
you can think of. This is the case here to: `st_bbox()`. 

```{r}
bbox <- st_bbox(meuse_sf)

bbox
```

The first and dirtiest method for plotting the data is just to use the 
`coord_sf()` function to provide limits using our `bbox` object.
There's nothing wrong with using this method, and we'll stick to it here,
but we note that a more elegant way might be to use `st_crop()`, which
works directly with our bounding box above.

```{r}
# Alternative solution: river_meuse <- st_crop(rivers, bbox)

ggplot(meuse_sf) +
  geom_sf(aes(size = lead)) +
  geom_sf(data = rivers) +
  coord_sf(
    xlim = c(bbox[["xmin"]], bbox[["xmax"]]),
    ylim = c(bbox[["ymin"]], bbox[["ymax"]])
  ) +
  theme(legend.position = "right")
```

Regrettably, this is not quite what we would like to see. The
resolution of this vector raster is too low and the river ends up
being just a line. Let's try something different.

# Raster Charts

What we'll do instead is to plot a raster map of the river and the surrounding
area. We'll download the map from stamen maps, using **ggmap**.

```{r}
library(ggmap)
```

To download a map, we need to call `get_stamenmap()` and specify a bounding
box. Fortunately for us, we already computed the bounding box of the points in
the last step, so let's use that.

```{r}
# extend the ranges slightly
xrng <- extendrange(c(bbox$xmin, bbox$xmax))
yrng <- extendrange(c(bbox$ymin, bbox$ymax))

meuse_raster <- get_stamenmap(
  c(
    left   = xrng[1],
    bottom = yrng[1],
    right  = xrng[2],
    top    = yrng[2]
  ),
  zoom = 14,
  maptype = "terrain"
)
```

To then build the map, we use the `ggmap()` function instead of `ggplot()`.
Now we run into a hurdle, however, because **ggmap** and **sf** unfortunately
does not play so well together^[In this case, the problem would actually
be hardly noticeable because the map is on a small scale and the problem,
which occurs because the bounding box of the raster is in a different projection,
results in no noticeable distorsion.]. As it stands, right now, it is
therefore better to use the standard map features when working with raster maps
from **ggmap**. To do so, we convert our meuse coordinates from 
simple features back to a standard data frame.

```{r, fig.cap = "Concentration of lead around Meuse.", fig.height = 6}
meuse_df <- cbind(
  st_drop_geometry(meuse_sf),
  st_coordinates(meuse_sf)
)
```

Now we can use the raster map together with our points, simply
overlaying them using the `geom_point()` function.

```{r}
ggmap(meuse_raster) +
  geom_point(aes(X, Y, size = lead), data = meuse_df) +
  theme(legend.position = "right")
```

# More Heavy Metal

Let's make the plot slightly more interesting by including data on the
other heavy metals too. One way to do so it to use facets, but for that to
work, we need to have our data in a long format, where one column denotes
the variable names and another notes their values.

```{r}
meuse_df_long <-
  meuse_df %>%
  pivot_longer(
    cadmium:zinc,
    names_to = "metal",
    values_to = "concentration"
  )
```

We construct our plot as before, faceting on the type of metal.

```{r, fig.cap = "Concentration of different heavy metals around the river Meuse in the Netherlands.", fig.width = 5, fig.height = 6}
ggmap(meuse_raster) +
  geom_point(
    aes(X, Y, size = concentration),
    alpha = 0.5,
    data = meuse_df_long
  ) +
  facet_wrap("metal") +
  theme(legend.position = "right")
```

# Geocoding

Purely for instructional reasons, let's assume that we were considering some
kind of cleanup of this heavy metal waste, but wanted to see how this would
affect affect the ruins of the nearby *Kasteel Stein* by displaying it on the
map.

We don't know the coordinates of the castle off-hand, so let's find them out by
geocoding. In the lecture, we used the facilities of **ggmap** for geocoding,
but unfortunately the only API that is currently supported in **ggmap** is
Google's API, which is not free.

Another alternative is to use
[tidygeocoder](https://CRAN.R-project.org/package=tidygeocoder), which features
a host of geocoding alternatives. A couple of these are free, for instance the
[Nominatim](https://nominatim.org/) API.^[The API is free, but it is also
rate-limited at one request per second and not as powerful as some of the other
available APIs. If you have a project where you need more serious geocoding,
you should probably get a key for one of the other APIs.]

Start by generating a `tibble` (or `data.frame`) with the names
and addresses that you want to geocode.

```{r}
library(tidygeocoder)

addresses <- tribble(
  ~ name,          ~ address,
  "Kasteel Stein", "Kasteel Stein, Netherlands"
)
```

Next, pipe these addresses on to `tidygeocoder::geocode()`.^[There is a
namespace clash here with **ggmap**, which has its own `ggmap::geocode()`
function, so we need to specify which function to use with `::` here.]

```{r}
kasteel_stein <-
  addresses %>%
  tidygeocoder::geocode(
    address,       # map to address
    method = "osm" # for Nominatim API
  ) 

kasteel_stein
```

To finish our map, we complement our previous plot with two layers for our new,
geocoded data, to show a label and a dot for the castle ruins.

```{r, fig.cap = "Concentration of different heavy metals around the river Meuse in the Netherlands.", fig.width = 6, fig.height = 7}
ggmap(meuse_raster) +
  geom_point(
    aes(X, Y, size = concentration),
    alpha = 0.5,
    data = meuse_df_long
  ) +
  geom_point(
    aes(long, lat),
    col = "red",
    data = kasteel_stein
  ) +
  geom_label(
    aes(long, lat, label = name),
    vjust = -0.5,
    data = kasteel_stein
  ) +
  facet_wrap("metal") +
  theme(legend.position = "right")
```

# Source Code

The source code for this document is available at
<https://github.com/stat-lu/dataviz/blob/main/worked-example-chile-plebiscite.Rmd>.
